---
title: "Code Red: Unpacking Inequality"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
```

```{r datacleaning, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rmarkdown::render("data_cleaning.rmd", quiet = TRUE)
```

```{r setup2, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Mapping set up
# Load the packages
library(sf)
library(leaflet)
library(tidycensus)

census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")
install = TRUE
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) |>
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts |>
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"  # This is a fallback in case there are other values
  ))

head(nyc_tracts)

```

```{r, echo=FALSE, warning=FALSE}
nyc_tracts = nyc_tracts |>
  rename(geoid = GEOID)

redlining_sf = 
  nyc_tracts |>
  left_join(redlining_snap, by = "geoid")

```

## Redlining Exploration:
Here is the map detailing the redlining grades given to areas in NYC. There are areas that our data set does not have information for, which are colored grey.

```{r redlining, echo=FALSE, warning=FALSE}

pal <- colorFactor(
  palette = c("#208e56",  
              "#e4e44c", 
              "#1595b4", 
              "#ef647b"),
  domain = redlining_sf$red_grade
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.9,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, "<br>Redlining Interval: ", interval2020, "<br>Redlining Score (2020): ", hrs2020)
  ) |>
  addLegend(
    pal = pal,
    values = ~red_grade,
    opacity = 0.7,
    title = "Redlining Category",
    position = "bottomright"
  )

```

### Distribution of Redlining Scores in Boroughs:

```{r, echo=FALSE, fig.align='center'}
redlining = redlining |>
  mutate(
    borough = case_when(
      fip == 36061 ~ "Manhattan",
      fip == 36047 ~ "Brooklyn",
      fip == 36081 ~ "Queens",
      fip == 36005 ~ "Bronx",
      fip == 36085 ~ "Staten Island",
      TRUE ~ "Unknown"))

  ggplot(redlining, aes(x = borough, y = hrs2020, fill = borough)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  coord_flip() +
   stat_summary(
     fun = mean, geom = "point", color = "blue", size = 3, shape = 18) +
  stat_summary(
    fun = median, geom = "point", color = "purple", size = 3, shape = 16) +
  labs(
    title = "Distribution of Redlining Scores by NYC Borough",
    x = "Borough",
    y = "Redlining Score",
    caption = "Mean (Blue Diamond), Median (Purple Circle)") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() 

```

This graph shows more detail. The distribution of the scores are fairly different for the boroughs, which Manhattan having the most distribution and Queens having the smallest.

### Redlining Grade Count by Borough:

```{r, echo=FALSE, fig.align='center'}
redlining |> 
  count(borough, red_grade) |>
  mutate(borough = fct_reorder(borough, n)) |>
  plot_ly(
    x = ~borough, 
    y = ~n, 
    color = ~red_grade, 
    type = "bar", 
    colors = "viridis"
  ) |>
  layout(
    title = "Count of Redlining Grades by Borough",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Count") 
  )
```

We can see here that Staten Island, Queens, and Bronx have the most C grade in their neighborhood. Manhattan and Brooklyn have the grade D as the most frequent. This is similar to what we saw in the violin plot previously. 

## SNAP Exploration and Demographics:

### SNAP Total % of Households within Redline Grade:

```{r, echo=FALSE, fig.align='center'}
redlining_snap |>
  separate(
    name,
    into = c("census_tract", "county", "state"),
    sep = "; ",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(
    borough = case_when(
      county == "Queens County" ~ "Queens",
      county == "Bronx County" ~ "Bronx",
      county == "New York County" ~ "Manhattan",
      county == "Kings County" ~ "Brooklyn",
      county == "Richmond County" ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  ) |>
  group_by(red_grade) |>
  summarize(
    avg_ph_snap = mean(ph_snap, na.rm = TRUE)
  ) |>
  ungroup() |>
  plot_ly(
    x = ~red_grade,
    y = ~avg_ph_snap,
    type = "bar",
    color = ~red_grade,
    colors = "viridis"
  ) |>
  layout(
    title = "Average SNAP Percentage of Households by Redline Grade in 2022",
    xaxis = list(title = "Redline Grade"),
    yaxis = list(title = "Average SNAP Percentage"),
    showlegend = FALSE
  )
  
```

### Race VS Redlinging Grade:

This is a stacked bar chart showing the race distribution of households by historical redlining grades (A, B, C, D, NA) in New York City.

```{r Race, echo=FALSE,warning=FALSE}
redlining_long <- redlining_sf |>
  select(red_grade, ph_white, ph_black, ph_asian, ph_other_race) |>
  pivot_longer(
    cols = c(ph_white, ph_black, ph_asian, ph_other_race),
    names_to = "race_group",
    values_to = "percent_household"
  ) |>
  mutate(race_group = case_when(
    race_group == "ph_white" ~ "White",
    race_group == "ph_black" ~ "Black",
    race_group == "ph_asian" ~ "Asian",
    race_group == "ph_other_race" ~ "Other"
  ))

ggplot(redlining_long, aes(x = red_grade, y = percent_household, fill = race_group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("White" = "#FFC0CB",   
                               "Black" = "#FFDDC1",   
                               "Asian" = "#BDECB6",   
                               "Other" = "#C5A3FF")) + 
  theme_minimal() +
  labs(
    title = "Race Distribution (White, Black, Asian, Other) by Redlining Grade",
    x = "Redlining Grade",
    y = "Percentage of Households",
    fill = "Race Group"
  ) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

* Grade D ("Hazardous") areas have a higher concentration of Black and minority households compared to Grade A or B areas, which show a higher proportion of White households.
* Grade C and D areas reflect the legacy of racial segregation, where redlined zones were predominantly occupied by communities of color.

## Healthy Grocery Store Exploration:

### Redlinging Grade VS. Grocery Stores(Manhattan,Brooklyn,and Bronx):

This map overlays the distribution of healthy grocery stores on top of the historical redlining grades across NYC.

```{r store1, echo=FALSE}
redlining_sf <- redlining_sf |>
  mutate(
    borough = case_when(
      str_detect(NAME, "Bronx County") ~ "Bronx",
      str_detect(NAME, "Kings County") ~ "Brooklyn",
      str_detect(NAME, "New York County") ~ "Manhattan",
      str_detect(NAME, "Queens County") ~ "Queens",
      str_detect(NAME, "Richmond County") ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  )

pal <- colorFactor(
  palette = c("#208e56",  
              "#e4e44c", 
              "#1595b4", 
              "#ef647b"),
  domain = redlining_sf$red_grade
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |>  
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>Redlining Grade: ", red_grade,
                    "<br>Borough: ", borough)
  ) |>
  addCircleMarkers(
    data = nyc_healthy_store,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "#0f6921",
    stroke = FALSE,
    fillOpacity = 0.7,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~red_grade,
    title = "Redlining Grade",
    opacity = 0.7
  ) |>
  addLegend(
    position = "bottomleft",
    colors = "#0f6921",
    labels = "Healthy Grocery Store",
    title = "Healthy Grocery Stores in NYC",
    opacity = 0.7
  )
```


### Redlinging Grade VS.Grocery Stores in Bronx Only:

This map zooms in on the Bronx area to focus on the distribution of healthy grocery stores within the context of historical redlining grades.

```{r bronx, echo=FALSE}
redlining_sf <- redlining_sf |>
  mutate(
    borough = case_when(
      str_detect(NAME, "Bronx County") ~ "Bronx",
      str_detect(NAME, "Kings County") ~ "Brooklyn",
      str_detect(NAME, "New York County") ~ "Manhattan",
      str_detect(NAME, "Queens County") ~ "Queens",
      str_detect(NAME, "Richmond County") ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  )

bronx_ <- redlining_sf |>
  filter(borough == "Bronx")

bronx_stores <- nyc_healthy_store |>
  filter(borough == "Bronx")


pal <- colorFactor(
  palette = c("#208e56",  
              "#e4e44c", 
              "#1595b4", 
              "#ef647b"),
  domain = redlining_sf$red_grade
)

leaflet(bronx_) |>
  addProviderTiles(providers$CartoDB.Positron) |>  
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>Redlining Grade: ", red_grade,
                    "<br>Borough: ", borough)
  ) |>
  addCircleMarkers(
    data = bronx_stores,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "#0f6921",
    stroke = FALSE,
    fillOpacity = 0.7,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~red_grade,
    title = "Redlining Grade",
    opacity = 0.7
  ) |>
  addLegend(
    position = "bottomleft",
    colors = "#0f6921",
    labels = "Healthy Grocery Store",
    title = "Healthy Grocery Stores in NYC",
    opacity = 0.7
  )
```

* Redlined areas (D) in the Bronx show fewer healthy grocery stores compared to blue (C) and yellow (B) areas.

## Health Outcome vs Redlining Scores

```{r, echo=FALSE, message=FALSE, fig.align='center'}
health_data =
  place_cleaned |>
  mutate(geoid = as.character(geoid)) |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()

health_data |>
  drop_na(red_grade, obesity, diabetes, highchol) |>
  select(red_grade, obesity, diabetes, highchol) |>
  group_by(red_grade) |>
  mutate(
    obesity_rate = mean(obesity, na.rm = FALSE),
    diabetes_rate = mean(diabetes, na.rm = FALSE),
    highcholesterol_rate = mean(highchol, na.rm = FALSE)
  ) |>
  select(red_grade, obesity_rate, diabetes_rate, highcholesterol_rate) |>
  pivot_longer(cols = c(obesity_rate, diabetes_rate, highcholesterol_rate), 
               names_to = "condition", values_to = "rate") |>
  ggplot(aes(x = condition, y = rate, fill = condition)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~red_grade) + 
  labs(
    title = "Diabetes, Obesity, and High Cholesterol Rates by Redlining Grade",
    x = "Condition",
    y = "Rate"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("diabetes_rate" = "#1f77b4",
                              "obesity_rate" = "#ff7f0e",  
                              "highcholesterol_rate" = "#2ca02c")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

```

The graph compares the relationship between redlining and health outcomes (diabetes, obesity, and high cholesterol) in New York City. The higher redlining scores (A, B) tend to have slightly higher obesity and high cholesterol rates, potentially indicating the complex intersection of socioeconomic status, healthcare access, and environmental factors related to historical redlining.


### SNAP Enrollment and Grocery Stores in Bronx(USE IN SHINY APP IF NEEDED).

```{r, echo=FALSE}
bronx_ <- redlining_sf |>
  filter(borough == "Bronx")

bronx_stores <- nyc_healthy_store |>
  filter(borough == "Bronx")

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = bronx_$ph_snap,
  na.color = "grey"
)

leaflet(bronx_) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~snap_pal(ph_snap),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>SNAP Enrollment (%): ", round(ph_snap, 2))
  ) |>
  addCircleMarkers(
    data = bronx_stores,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,  
    color = "#0f6921", 
    stroke = FALSE,
    fillOpacity = 0.7,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    pal = snap_pal,
    values = ~ph_snap,
    opacity = 0.7,
    title = "SNAP Enrollment (%)",
    position = "bottomright"
  ) |>
  addLegend(
    position = "bottomleft",
    colors = "#0f6921",
    labels = "Healthy Grocery Store",
    title = "Healthy Grocery Stores in Bronx",
    opacity = 0.7
  )

```

