---
title: "Code Red: Unpacking Inequality"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
```

```{r datacleaning, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rmarkdown::render("data_cleaning.rmd", quiet = TRUE)
```

```{r setup2, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Mapping set up
# Load the packages
library(sf)
library(leaflet)
library(tidycensus)

census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")
install = TRUE
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) |>
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts |>
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"  # This is a fallback in case there are other values
  ))

head(nyc_tracts)

```

```{r, echo=FALSE, warning=FALSE}
nyc_tracts = nyc_tracts |>
  rename(geoid = GEOID)

redlining_sf = 
  nyc_tracts |>
  left_join(redlining_snap, by = "geoid")

```

## Redlining Exploration:

### Redlining Throughout Boroughs:

Here is the map detailing the redlining scores (ranging from 1-4) given to areas in NYC. There are areas that our data set does not have information for, which are colored grey.

```{r, echo=FALSE, warning=FALSE}

borough_boundaries <- nyc_tracts |>
  group_by(borough) |>
  summarize(geometry = st_union(geometry), .groups = 'drop')

pal <- colorNumeric(
  palette = "YlOrRd",  
  domain = redlining_sf$hrs2020, 
  na.color = "grey"  
)

leaflet(redlining_sf) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(hrs2020),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>Borough: ", borough, 
                    "<br>Redlining Score (2020): ", hrs2020)
  ) |>
  addPolygons(
    data = borough_boundaries,
    weight = 3,
    color = "black",
    fill = FALSE,
    opacity = 1,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#000000",
      bringToFront = TRUE
    )
  ) |>
  addLabelOnlyMarkers(
    data = st_centroid(borough_boundaries),
    lng = ~st_coordinates(geometry)[,1],
    lat = ~st_coordinates(geometry)[,2],
    label = ~borough,
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = 'center',
      textOnly = TRUE,
      style = list("color" = "black", "font-size" = "16px", "font-weight" = "bold")
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~hrs2020,
    opacity = 0.7,
    title = "Redlining Score (2020)",
    position = "bottomright"
  )

```

### Average Redlining Score in Each Borough:

Lets look at the average redlining score in each borough: 

```{r, echo=FALSE, warning=FALSE}
redlining = redlining |>
  mutate(
    borough = case_when(
      fip == 36061 ~ "Manhattan",
      fip == 36047 ~ "Brooklyn",
      fip == 36081 ~ "Queens",
      fip == 36005 ~ "Bronx",
      fip == 36085 ~ "Staten Island",
      TRUE ~ "Unknown")
  )

redlining |>
  group_by(borough) |>
  summarize(avg_redline_score =
              mean(hrs2020, na.rm = TRUE)) |>
  mutate(borough = fct_reorder(borough, avg_redline_score)) |>
  ggplot(aes(x = borough, y = avg_redline_score, fill = borough)) +
  geom_point() +
  labs(
    title = "Average Redlining Score by Borough",
    x = "Borough",
    y = "Average Redlining Score"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

This graph shows the average redlining scores in each borough. Although it looks fairly different, the range is fairly low, between ~3.06 and ~3.24. This does not give us too much information. Lets look at the distribution: 

### Distribution of Redlining Scores in Boroughs:

```{r, echo=FALSE}
redlining |>
  ggplot(aes(x = borough, y = hrs2020, fill = borough)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  labs(
    title = "Distribution of Redlining Scores by NYC Borough",
    x = "Borough",
    y = "Redlining Score"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

This graph shows more detail. The distribution of the scores are fairly different for the boroughs, which Manhattan having the most distribution and Queens having the smallest.

### Redlining Grade Count by Borough:

```{r, echo=FALSE}
redlining |> 
  count(borough, red_grade) |>
  mutate(borough = fct_reorder(borough, n)) |>
  plot_ly(
    x = ~borough, 
    y = ~n, 
    color = ~red_grade, 
    type = "bar", 
    colors = "viridis"
  ) |>
  layout(
    title = "Count of Redlining Grades by Borough",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Count") 
  )

```

We can see here that Staten Island, Queens, and Bronx have the most C grade in their neighborhood. Manhattan and Brooklyn have the grade D as the most frequent. This is similar to what we saw in the violin plot previously. 

## SNAP Exploration and Demographics:

### SNAP Total Household Count in Boroughs:

```{r, echo=FALSE}
SNAP = SNAP |>
  separate(
    name,
    into = c("census_tract", "county", "state"),
    sep = "; ",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(
    borough = case_when(
      county == "Queens County" ~ "Queens",
      county == "Bronx County" ~ "Bronx",
      county == "New York County" ~ "Manhattan",
      county == "Kings County" ~ "Brooklyn",
      county == "Richmond County" ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  )

SNAP |>
  group_by(borough) |>
  summarize(
    total_household_count = sum(total_ct_households, na.rm = TRUE)
  ) |>
  filter(borough != "Unknown") |>
  plot_ly(
    x = ~borough, 
    y = ~total_household_count, 
    color = ~borough, 
    type = "bar", 
    colors = "viridis"
  ) |>
  layout(
    title = "Total SNAP-Enrolled Households by Borough in 2022",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Total Household Count"),
    showlegend = FALSE
  )
```

### SNAP Enrollment Demographics (RACE):

```{r, echo=FALSE, warning=FALSE}
borough_snap_map = redlining_sf |>
  separate(
    name,
    into = c("census_tract", "county", "state"),
    sep = "; ",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(
    borough = case_when(
      county == "Queens County" ~ "Queens",
      county == "Bronx County" ~ "Bronx",
      county == "New York County" ~ "Manhattan",
      county == "Kings County" ~ "Brooklyn",
      county == "Richmond County" ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  ) |>
  group_by(borough)
```

```{r, echo=FALSE, warning=FALSE}
race_snap_data <- borough_snap_map |>
  select(borough, ph_white, ph_black, ph_asian, ph_hispanic, ph_other_race, ph_snap) |>
  group_by(borough) |>
  summarise(
    avg_white = mean(ph_white, na.rm = TRUE),
    avg_black = mean(ph_black, na.rm = TRUE),
    avg_asian = mean(ph_asian, na.rm = TRUE),
    avg_hispanic = mean(ph_hispanic, na.rm = TRUE),
    avg_other_race = mean(ph_other_race, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = starts_with("avg"),
    names_to = "race",
    values_to = "average_percentage"
  ) |>
  mutate(
    race = case_when(
      race == "avg_white" ~ "White",
      race == "avg_black" ~ "Black",
      race == "avg_asian" ~ "Asian",
      race == "avg_hispanic" ~ "Hispanic",
      race == "avg_other_race" ~ "Other Race",
      TRUE ~ race
    )
  )

ggplot(race_snap_data, aes(x = borough, y = average_percentage, fill = race)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(
    title = "Distribution of Race Among SNAP Enrollment by Borough",
    x = "Borough",
    y = "Proportion of Households (%)",
    fill = "Race"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```

## Healthy Grocery Store Exploration:

### Grocery Stores in Manhattan, Brooklyn, and Bronx:

```{r, echo=FALSE}
leaflet() |>
  addTiles() |>
  addCircleMarkers(
    data = nyc_healthy_store,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    position = "bottomright",
    colors = "blue",
    labels = "Grocery Store",
    title = "Healthy Grocery Stores"
  )

```

### Grocery Stores in Bronx Only:

```{r, echo=FALSE}
bronx_stores <- nyc_healthy_store |>
  filter(borough == "Bronx")

leaflet() |>
  addTiles() |>
  addCircleMarkers(
    data = bronx_stores,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    position = "bottomright",
    colors = "blue",
    labels = "Grocery Store",
    title = "Healthy Grocery Stores in Bronx"
  )

```

## Health: Obesity vs Redlining Scores

```{r, echo=FALSE, message=FALSE}
health_data =
  place_cleaned |>
  mutate(geoid = as.character(geoid)) |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names() |>
  mutate(
    borough = case_when(
      fip == 36061 ~ "Manhattan",
      fip == 36047 ~ "Brooklyn",
      fip == 36081 ~ "Queens",
      fip == 36005 ~ "Bronx",
      fip == 36085 ~ "Staten Island",
      TRUE ~ "Unknown")
  )

health_data |>
  filter(!is.na(obesity) & !is.na(hrs2020)) |>
  ggplot(aes(y = obesity, x = hrs2020, color = borough)) +
  geom_point() +
  geom_smooth(color = "black") +
  facet_wrap(~borough, scales = "free") +
  labs(
    title = "Obesity Rate vs Redlining Score",
    x = "Redlining Score (HRS 2020)",
    y = "Obesity Rate (%)"
  ) +
  theme_minimal()
```

The exploratory analysis of obesity rates against redlining scores reveals a consistent positive trend across most of New York City's boroughs, with areas of historical redlining showing higher obesity rates. While Manhattan presents a weaker relationship, other boroughs like the Bronx, Brooklyn, Queens, and Staten Island demonstrate stronger associations, suggesting that the legacy of redlining may contribute to health disparities. 

## Exploration of Redlining, SNAP, and Healthy Grocery Stores

### Redlining Scores and Demographics:

```{r, echo=FALSE}
pal <- colorFactor(
  palette = c("green", "blue", "yellow", "red"),
  domain = redlining_sf$red_grade
)

leaflet(redlining_sf) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, "<br>Redlining Interval: ", interval2020, "<br>Redlining Score (2020): ", hrs2020)
  ) |>
  addLegend(
    pal = pal,
    values = ~red_grade,
    opacity = 0.7,
    title = "Redlining Category",
    position = "bottomright"
  )
```

### Redlining Scores and SNAP Enrollment.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
borough_snap_map |>
  ggplot(aes(x = hrs2020, y = ph_snap, color = borough)) +
  geom_jitter(alpha = 0.6, size = 2, width = 0.1, height = 1) + 
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  facet_wrap(~ borough, scales = "free") +
  labs(
    title = "Correlation Between Redlining Scores and SNAP Enrollment Rates by Borough",
    subtitle = "Analysis of SNAP enrollment rates relative to redlining scores (HRS 2020)",
    x = "Redlining Score (HRS 2020)",
    y = "SNAP Enrollment (%)",
    color = "Borough"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom"
  )

```

### Redlining and Grocery Stores in Bronx

```{r, echo=FALSE}
redlining_sf <- redlining_sf |>
  mutate(
    borough = case_when(
      str_detect(NAME, "Bronx County") ~ "Bronx",
      str_detect(NAME, "Kings County") ~ "Brooklyn",
      str_detect(NAME, "New York County") ~ "Manhattan",
      str_detect(NAME, "Queens County") ~ "Queens",
      str_detect(NAME, "Richmond County") ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  )

bronx_redlining <- redlining_sf |>
  filter(borough == "Bronx")

bronx_stores <- nyc_healthy_store |>
  filter(borough == "Bronx")

pal <- colorFactor(
  palette = c("green", "blue", "yellow", "red"),
  domain = bronx_redlining$red_grade
)

leaflet(bronx_redlining) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>Redlining Grade: ", red_grade)
  ) |>
  addCircleMarkers(
    data = bronx_stores,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    pal = pal,
    values = ~red_grade,
    opacity = 0.7,
    title = "Redlining Grade",
    position = "bottomright"
  ) |>
  addLegend(
    position = "bottomleft",
    colors = "blue",
    labels = "Grocery Store",
    title = "Healthy Grocery Stores in Bronx"
  )

```

### SNAP Enrollment and Grocery Stores in Bronx.

```{r, echo=FALSE}
bronx_ <- borough_snap_map |>
  filter(borough == "Bronx")

bronx_stores <- nyc_healthy_store |>
  filter(borough == "Bronx")

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = bronx_$ph_snap,
  na.color = "grey"
)

leaflet(bronx_) |>
  addTiles() |>
  addPolygons(
    fillColor = ~snap_pal(ph_snap),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>SNAP Enrollment (%): ", round(ph_snap, 2))
  ) |>
  addCircleMarkers(
    data = bronx_stores,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.8,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>
  addLegend(
    pal = snap_pal,
    values = ~ph_snap,
    opacity = 0.7,
    title = "SNAP Enrollment (%)",
    position = "bottomright"
  ) |>
  addLegend(
    position = "bottomleft",
    colors = "blue",
    labels = "Grocery Store",
    title = "Healthy Grocery Stores in Bronx"
  )

```


