---
title: "Code Red: Unpacking Inequality"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
```

```{r datacleaning, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rmarkdown::render("data_cleaning.rmd", quiet = TRUE)
```

```{r setup2, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Mapping set up
# Load the packages
library(sf)
library(leaflet)
library(tidycensus)

census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")
install = TRUE
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) |>
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts |>
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"  # This is a fallback in case there are other values
  ))

head(nyc_tracts)

```

```{r, echo=FALSE, warning=FALSE}
nyc_tracts = nyc_tracts |>
  rename(geoid = GEOID)

redlining_sf = 
  nyc_tracts |>
  left_join(redlining_snap, by = "geoid")

```

## Redlining Exploration

Below is the map detailing the redlining grades given to areas in NYC. The corresponding colors and grades are in the legend. There are areas that our data set does not have information for, which are colored grey.

```{r redlining, echo=FALSE, warning=FALSE}

pal <- colorFactor(
  palette = c("#208e56",  
              "#e4e44c", 
              "#1595b4", 
              "#ef647b"),
  domain = redlining_sf$red_grade
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.9,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, "<br>Redlining Interval: ", interval2020, "<br>Redlining Score (2020): ", hrs2020)
  ) |>
  addLegend(
    pal = pal,
    values = ~red_grade,
    opacity = 0.7,
    title = "Redlining Category",
    position = "bottomright"
  )

```
This map effectively captures the historical spatial boundaries of redlining across NYC boroughs, showing how systemic inequities were distributed geographically.

- Manhattan and Brooklyn show stark contrasts between privileged (green/blue) and marginalized (yellow/red) areas.
- The Bronx and Brooklyn have the highest concentration of historically "hazardous" (D) zones, reflecting long-standing inequities.
- Queens and Staten Island had relatively fewer redlined areas, suggesting different historical patterns of urban development and disinvestment.

### Distribution of Redlining Scores in Boroughs

Lets look at Redlining Scores in a different way

```{r, echo=FALSE, fig.align='center'}
redlining = redlining |>
  mutate(
    borough = case_when(
      fip == 36061 ~ "Manhattan",
      fip == 36047 ~ "Brooklyn",
      fip == 36081 ~ "Queens",
      fip == 36005 ~ "Bronx",
      fip == 36085 ~ "Staten Island",
      TRUE ~ "Unknown"))

  ggplot(redlining, aes(x = borough, y = hrs2020, fill = borough)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  coord_flip() +
   stat_summary(
     fun = mean, geom = "point", color = "blue", size = 3, shape = 18) +
  stat_summary(
    fun = median, geom = "point", color = "purple", size = 3, shape = 16) +
  labs(
    title = "Distribution of Redlining Scores by NYC Borough",
    x = "Borough",
    y = "Redlining Score",
    caption = "Mean (Blue Diamond), Median (Purple Circle)") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() 

```

This graph shows different details. The distribution of the scores are fairly different for each of the boroughs, with Manhattan having the largest range and Queens having the smallest. Additionally, the mean and median redlining score for each of the borough are almost overlapping/the same. However, Manhattan has pretty different mean and median (mean closer to 3 and median closer to 4).

### Redlining Grade Count by Borough

```{r, echo=FALSE, fig.align='center'}
set2_palette = RColorBrewer::brewer.pal(n = 4, name = "Set2") 

redlining |> 
  count(borough, red_grade) |> 
  mutate(borough = fct_reorder(borough, n)) |> 
  plot_ly(
    x = ~borough,
    y = ~n,
    color = ~red_grade,
    type = "bar",
    colors = set2_palette
  ) |> 
  layout(
    title = "Count of Redlining Grades by Borough",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Count"),
    barmode = "stack" 
  )
```

This graph shows the grade count in each of the boroughs. Staten Island, Queens, and Bronx have the most C grade in their neighborhood (followed by D grade in second place). Manhattan and Brooklyn have grade D as the most frequent. This is reflective of what was previously observed in the violin plot. 

## SNAP Exploration and Demographics

### SNAP Enrollment

```{r, echo=FALSE, warning=FALSE}

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = redlining_sf$ph_snap,
  na.color = "grey"
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~snap_pal(ph_snap),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>SNAP Enrollment (%): ", round(ph_snap, 2))
  ) |>

  addLegend(
    pal = snap_pal,
    values = ~ph_snap,
    opacity = 0.7,
    title = "SNAP Enrollment (%)",
    position = "bottomright"
  ) 
```
This map illustrates the strong correlation between historical redlining and present-day SNAP enrollment, underscoring how systemic inequities continue to influence food assistance needs across NYC.

* The Bronx stands out with the highest overall SNAP enrollment, aligning with its high concentration of historically redlined zones and lower-income households.
* Brooklyn also shows significant SNAP participation in neighborhoods historically marked as “hazardous.”
* Queens and Staten Island have lower overall SNAP participation, which reflects fewer redlined areas and more suburban development patterns.


### SNAP Total % of Households within Redline Grade

```{r, echo=FALSE, fig.align='center'}
redlining_snap |>
  separate(
    name,
    into = c("census_tract", "county", "state"),
    sep = "; ",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(
    borough = case_when(
      county == "Queens County" ~ "Queens",
      county == "Bronx County" ~ "Bronx",
      county == "New York County" ~ "Manhattan",
      county == "Kings County" ~ "Brooklyn",
      county == "Richmond County" ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  ) |>
  group_by(red_grade) |>
  summarize(
    avg_ph_snap = mean(ph_snap, na.rm = TRUE)
  ) |>
  ungroup() |>
  plot_ly(
    x = ~red_grade,
    y = ~avg_ph_snap,
    type = "bar",
    color = ~red_grade,
    colors = set2_palette
  ) |>
  layout(
    title = "Average SNAP Percentage of Households by Redline Grade in 2022",
    xaxis = list(title = "Redline Grade"),
    yaxis = list(title = "Average SNAP Percentage"),
    showlegend = FALSE
  )
  
```

This graph depicts the average % of households on SNAP within the areas with the grades. Unsurprisingly, areas with redlining grade of D have the highest average % of households on SNAP (23.7%), followed by C (18.6%), then B (17.2) and A (5.7) with the lowest. 

### Race VS Redlinging Grade

This is a stacked bar chart showing the race distribution of households by historical redlining grades (A, B, C, D, NA) in New York City.

```{r Race, echo=FALSE,warning=FALSE}
redlining_long <- redlining_sf |>
  select(red_grade, ph_white, ph_black, ph_asian, ph_other_race) |>
  group_by(red_grade)|>
  summarise(pha_white = mean(ph_white, na.rm = TRUE),
            pha_black = mean(ph_black, na.rm = TRUE),
            pha_asian = mean(ph_asian, na.rm = TRUE),
            pha_other_race = mean(ph_other_race, na.rm = TRUE))|>
  pivot_longer(
    cols = c(pha_white, pha_black, pha_asian, pha_other_race),
    names_to = "race_group",
    values_to = "percent_household"
  ) |>
  mutate(race_group = case_when(
    race_group == "pha_white" ~ "White",
    race_group == "pha_black" ~ "Black",
    race_group == "pha_asian" ~ "Asian",
    race_group == "pha_other_race" ~ "Other"
  ))

ggplot(redlining_long, aes(x = red_grade, y = percent_household, fill = race_group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set2") + 
  theme_minimal() +
  labs(
    title = "Race Distribution (White, Black, Asian, Other) by Redlining Grade",
    x = "Redlining Grade",
    y = "Percentage of Households",
    fill = "Race Group"
  ) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

* Grade D ("Hazardous") areas have a higher concentration of Black and minority households compared to Grade A or B areas, which show a higher proportion of White households.
* Grade C and D areas reflect the legacy of racial segregation, where redlined zones were predominantly occupied by communities of color.

## Healthy Grocery Store Exploration

### Grocery Stores (Manhattan, Brooklyn and Bronx)

```{r, echo=FALSE, warning=FALSE}

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = redlining_sf$ph_snap,
  na.color = "grey"
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(
    data = nyc_healthy_store,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,  
    color = "#0f6921", 
    stroke = FALSE,
    fillOpacity = 0.7,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>

  addLegend(
    position = "bottomleft",
    colors = "#0f6921",
    labels = "Healthy Grocery Store",
    title = "Healthy Grocery Stores in Bronx",
    opacity = 0.7
  )

```
This map provide a focused exploration of healthy grocery stores in three boroughs: Manhattan, Brooklyn, and the Bronx.
* The density of healthy grocery stores in Bronx is uneven across neighborhoods, with higher concentrations in specific areas.
* In Brooklyn, healthy grocery stores are relatively well-distributed, but there are clear clusters in central and northern neighborhoods. Southern Brooklyn appears less represented in this dataset, potentially reflecting disparities in access.


## Health Outcomes Exploration

### Health Outcome vs Redlining Scores

```{r, echo=FALSE, message=FALSE, fig.align='center'}
health_data =
  place_cleaned |>
  mutate(geoid = as.character(geoid)) |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()

health_data |>
  drop_na(red_grade, obesity, diabetes, highchol) |>
  select(red_grade, obesity, diabetes, highchol) |>
  group_by(red_grade) |>
  mutate(
    obesity_rate = mean(obesity, na.rm = FALSE),
    diabetes_rate = mean(diabetes, na.rm = FALSE),
    highcholesterol_rate = mean(highchol, na.rm = FALSE)
  ) |>
  select(red_grade, obesity_rate, diabetes_rate, highcholesterol_rate) |>
  pivot_longer(cols = c(obesity_rate, diabetes_rate, highcholesterol_rate), 
               names_to = "condition", values_to = "rate") |>
  ggplot(aes(x = condition, y = rate, fill = condition)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~red_grade) + 
  labs(
    title = "Diabetes, Obesity, and High Cholesterol Rates by Redlining Grade",
    x = "Condition",
    y = "Rate"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

```

The graph shows how redlining grades correlate with health outcomes (diabetes, obesity, and high cholesterol) in New York City. The neighborhoods with A grades have slightly lower rates of diabetes and obesity, but D areas suffer from higher rates of these conditions. High cholesterol rates are interestingly higher in better-graded areas, suggesting that factors such as access to healthcare and preventive services may play a significant role in these health outcomes.



