---
title: "Final Report"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Motivation

The practice of redlining, where certain neighborhoods were systematically denied access to resources based on racial and economic factors, has had long-lasting effects on urban communities. In New York City, the legacy of redlining continues to shape access to vital services such as healthy food and public assistance programs. The motivation for this project is to explore how these historical practices have impacted modern-day disparities in access to healthy grocery stores and enrollment in the Supplemental Nutrition Assistance Program (SNAP).

Our primary goal is to investigate the relationship between the neighborhoods that were historically redlined and their current access to essential resources. We seek to answer questions about the extent to which redlining has contributed to food deserts, areas where healthy food options are scarce, and whether these communities have disproportionately higher rates of SNAP enrollment. By understanding these connections, we aim to uncover the ways in which systemic inequalities persist in shaping the health and well-being of New Yorkers, and to contribute to ongoing discussions about potential policy solutions to address these disparities.

This project is driven by a desire to shed light on the intersection of urban planning, socioeconomic disparities, and public health, and to highlight the role of policy in creating both opportunities and barriers for communities that continue to experience the effects of redlining. Through this exploration, we hope to contribute valuable insights to the broader conversation about equity in urban environments and the long-term impacts of discriminatory practices on marginalized populations.

# Initial Questions

- **Neighborhood:** What areas in NYC were most heavily impacted by redlining? How do the demograpics within these areas compare?
- **Grocery:** Are there fewer healthy grocery stores in neighborhoods that were historically redlined compared to other areas? How does access to health grocery stores correlate with health outcomes in these neighborhoods?
- **SNAP:** Is there a higher rate of SNAP enrollment in areas that were historically redlined? Do these areas face more challenges accessing SNAP benefits? 
- **Analysis:**
  - How does redlining classification (e.g., grade A, B, C, D) predict the proportion of households enrolled in SNAP at the census tract level?
  - Is redlining associated with food insecurity related health outcomes(diabetes, obesity and high cholesterol level)?

# Data

## Data Sources 

### SNAP enrollment data:

The SNAP enrollment data was acquired from the American Community Survey 5-year estimates of 2022 from the Census Bureau. The original dataset contained estimates for SNAP/Food stamp enrollment at the census tract level as well as other households attributes at the census tract level: estimate and percent for distribution of race and ethnicity, poverty status, disability status, presentation of children under age of 18, etc.

We selected the variables that were relevant to our analysis to build the final dataset, the final variables included were the following:
  
  - ` r geoid`: the geographic id for identification of census tract 
  - `total_ct_households`: estimate total count for number of households 
  - `ph_elderly`: estimate percentage of households with one or more people in the household to be 60 or over
  - `ph_with_children`: estimate percentage of households with children under 18 years
  - `ph_no_children`: estimate percentage of households without children under 18 years 
  - `ph_below_poverty`:estimate percentage of households below poverty level in the past 12 months
  - `ph_above_poverty`: estimate percentage of households at or above poverty level in the past 12 months
  - `ph_disability`: estimate percentage of households with one or more people with disability 
  - `ph_no_disability`: estimate percentage of households with no person with disability 
  - `ph_white`: estimate percent households race and Hispanic or Latino origin of householder, White alone
  - `ph_black`: estimate percent households race and Hispanic or Latino origin of householder, Black or African American alone
  - `ph_asian`: estimate percent households race and Hispanic or Latino origin of householder, Asian alone
  - `ph_other race`: estimate percent households race and Hispanic or Latino origin of householder, other races (derived from the estimate value for White, Black and Asian race groups)
  - `ph_hispanic`: estimate percent households race and Hispanic or Latino origin of householder, Hispanic or Latino origin of any race
  - `ph_non_hispanic`: estimate percent households race and Hispanic or Latino origin of householder, not of Hispanic or Latino origin of any race  (derived from the estimate value for Hispanic or Latino origin of any race)
  - `ph_no_work`:estimate percent of families with no workers in past 12 months 
  - `ph_1_work`: estimate percent of families with 1 worker in past 12 months 
  - `ph_2_work`: estimate percent of families with 2 or more workers in past 12 months 
  - `ph_snap`: estimate percent of households receiving SNAP/Food Stamp 

### Historic Redlining Score Data

The historic redlining score (HRS) (`hrs2020`) was obtained from the calculation by the National Community Reinvestment Coalition(NCRC). It was a weighted score calculated based on the percentage of the area within each 2020 census tract according to HOLC grades. HRS ranged from 1-4, the higher the score indicated a greater level of historical redlining. The detailed method for calculating the HRS was indicated in the figure below. The hrs data was divided into quartiles and the division of quartiles is indicated by the `interval2020` variable. The more detailed methodology regarding coding for redlining grades can be found [here](https://ncrc.org/holc-health/#methods).

We filtered based on the geoid to only keep the data for the census tracts within NYC based on the fipcodes for NYC counties. We recoded the census tracts with the Home Owners’ Loan Corporation (HOLC) grades according to HRS. A (HRS: 1~1.5) defined as "Best", B (HRS: 1.5~2.5) defined as "Still Desirable", C (HRS: 2.5~3.5) defined as "Definitely Declining” and D(HRS > 3.5) defined as "Hazardous”.

## Recognized Healthy Store in NYC 

We obtained a data regarding recognized healthy store from the NYC Department of Health and Mental Hygiene. This dataset contained 675 bodegas & grocery stores receiving recognition from Borough President's Office regarding providing healthy food choices to its customers. The original dataset contained variables describing basic information of the stores(`name`,`store identifier`), geographic location(`address`, `census tract`, `zip code`, `borough`, `longitude` and `latitude`), year awarded and the program wave. 

## PLACE data 

We also extracted the PLACES: Local Data for Better Health, 2024 release for NYC census tracts from CDC to obtain the prevalence of a variety of health indicators that are likely associated with food insecurity: `diabetes`, `obesity` and `high cholesterol`. 

## Data Cleaning

```{r overall_setup, message=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
```

### Cleaning SNAP data set: 

```{r}
SNAP = 
  read_csv(file = "./Datasets/ACS_SNAP_ENROLLMENT.csv")|>
  janitor::clean_names()|>
  select(
    geo_id:s2201_c01_001e,
    s2201_c02_002e,
    s2201_c02_009e,
    s2201_c02_015e,
    s2201_c02_021e:s2201_c02_028e,
    s2201_c02_032e,
    s2201_c02_036e:s2201_c02_038e,
    s2201_c04_001e)|>
  select(-ends_with("m"))|>
  slice(-1)|>
  mutate(geoid = str_remove(geo_id, ".*US"))|>
  rename(
     total_ct_households = s2201_c01_001e,
     ph_elderly = s2201_c02_002e,
     ph_with_children = s2201_c02_009e,
     ph_no_children = s2201_c02_015e,
     ph_below_poverty = s2201_c02_021e,
     ph_above_poverty = s2201_c02_022e,
     ph_disability = s2201_c02_023e,
     ph_no_disability = s2201_c02_024e,
     ph_white = s2201_c02_025e,
     ph_black = s2201_c02_026e,
     ph_asian = s2201_c02_028e,
     ph_hispanic = s2201_c02_032e,
     ph_no_work = s2201_c02_036e,
     ph_1_work = s2201_c02_037e,
     ph_2_work = s2201_c02_038e,
     ph_snap = s2201_c04_001e
  )|>
  mutate(across(total_ct_households:ph_snap, as.numeric),
         ph_other_race = 100-ph_white-ph_black-ph_asian,
         ph_non_hispanic = 100-ph_hispanic)|>
  select(geoid,name,ph_snap,everything())|>
  select(-geo_id)
```

The window below is a preview of the SNAP data set. 

```{r}
head(SNAP)|>
  knitr::kable()|>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

### Redlining dataset cleaning:

```{r}
redlining = 
  read_excel("./Datasets/Historic Redlining Score 2020B.xlsx")|>
  janitor::clean_names()|>
  select(-cbsa)|>
  mutate(fip = substr(geoid20, 1, 5),
         interval2020 = as.character(interval2020),
         red_grade = case_when(
           hrs2020 < 1.5 ~ "A",
           hrs2020 < 2.5 ~ "B",
           hrs2020 < 3.5 ~ "C",
           TRUE ~ "D"
         ))|>
  filter(fip %in% c("36005", "36047", "36061", "36081", "36085"))|>
  rename(geoid = geoid20)
```

We subsequently match the historic redlining score with snap enrollment data and our analysis based on geoid for the census tracts. 18 census tracts were removed for the final analysis due to missing value for snap enrollment data. Our final dataset contained 2034 observations and 25 variables. 

```{r}
redlining_snap = 
  redlining|>
  left_join(SNAP, by = "geoid")|>
  filter(is.na(ph_snap)==FALSE)

redlining_snap|>
  knitr::kable()|>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

### Health stores data set cleaning:

```{r}
nyc_healthy_store = 
  read_csv("./Datasets/Recognized_Shop_Healthy_Stores.csv")|>
  janitor::clean_names()|>
  distinct(bin, .keep_all = TRUE)|>
  mutate(fipcode = case_when(
    borough == "Bronx" ~ "36005",
    borough == "Brooklyn" ~ "36047",
    borough == "New York" ~ "36061"
  ),
  ct_label = case_when(
    census_tract_2020 < 100 ~ paste0("00", census_tract_2020, "00"),
    census_tract_2020 < 1000 ~ paste0("0", census_tract_2020, "00"),
    census_tract_2020 < 1200 ~ paste0(census_tract_2020, "00"),
    census_tract_2020 < 10000 ~ paste0("00", census_tract_2020),
    census_tract_2020 < 100000 ~ paste0("0", census_tract_2020)
  ))|>
  select(store_name,borough,zip_code,latitude,longitude,fipcode, ct_label)|>
  filter(!is.na(ct_label))|>
  mutate(geoid = paste(fipcode, ct_label, sep = ""))|>
  select(-fipcode, -ct_label)

head(nyc_healthy_store)|>
  knitr::kable()|>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

The data only had stores from the Bronx, Brooklyn and New York County. We coded the geoid for the census tracts based on the fipcode of NYC, boroughs and specific census tract codes. Stores that does not provide the census tract level data were excluded.Eventually, we selected the variables that were relevant for our exploratory and statistical analysis. The variables included involve: `r names(nyc_healthy_store)`.

We obtain the count for the number of healthy food stores within each census tracts and matched that with the snap enrollment data for New York City, Bronx and Brooklyn Borough based on geoid.  

```{r}
health_store_count = 
nyc_healthy_store|>
  group_by(geoid)|>
  summarise(count_healthy_stores=n())

#please note that healthy store count is only available in three boroughs so make sure you filter before the match
#example code for the match

redlining_snap|>
  filter(str_starts(geoid, "36005") | str_starts(geoid, "36047")| str_starts(geoid, "36061"))|>
  left_join(health_store_count,by="geoid")
```

### PLACE data 

For the purpose of matching with other data sets, we pivoted the original data set into wide format to obtain the health measures as columns and rows as each census tracts. 

```{r}
place_crude =
  read_csv("./Datasets/PLACES.csv")|>
  janitor::clean_names()|>
  rename(geoid = location_id)|>
  select(geoid, measure:data_value, measure_id)


place_cleaned=
  place_crude|>
  filter(measure_id %in% c("DIABETES", "HIGHCHOL", "OBESITY"))|>
  pivot_wider(
    id_cols = geoid,
    names_from = measure_id,
    values_from = data_value
  )|>
  janitor::clean_names()|>
  mutate(geoid = as.character(geoid))

head(place_cleaned)|>
  knitr::kable()|>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

The description regarding the health indicators we have used for the analysis were indicated in the table below. 

```{r}
place_crude|>
  distinct(measure, data_value_unit, measure_id)|>
  filter(measure_id %in% c("DIABETES", "HIGHCHOL", "OBESITY"))|>
  knitr::kable()
```

# Exploratory Analysis and Visualizations

Exploratory analysis was a critical part of this project, enabling us to uncover patterns, trends, and relationships within the data. By visualizing redlining grades, SNAP enrollment, demographic distributions, and health outcomes, we identified significant disparities and connections between historical policies and current realities. The analysis revealed correlations which helped us refine our hypotheses and focus on targeted analyses, including borough-level comparisons and the impact of redlining on health disparities.

Map visualizations play a key role in this project by bringing the data to life and showing the connections between historical redlining, access to healthy grocery stores, and SNAP enrollment. By presenting complex data in a geographic context, these visualizations facilitate the identification of patterns and disparities across neighborhoods, highlighting the enduring impacts of systemic inequities. With this method, we want to promote informed conversations and analysis of possible pathways for improved equality in New York City by offering a clear and fact-based knowledge of the long-term effects of discriminatory policies.

## Data Preparation and Spatial Integration

We loaded essential libraries for data manipulation, visualization, and interactive plots, and we pulled ACS data for NYC census tracts using tidycensus, assigned borough names based on county names, transformed the spatial data for compatibility with leaflet, and merged this spatial data with redlining and SNAP-related datasets for further analysis and visualization.

```{r setup, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
```

```{r datacleaning, results='hide', message=FALSE, warning=FALSE}
rmarkdown::render("data_cleaning.rmd", quiet = TRUE)
```

```{r setup2, results='hide', message=FALSE, warning=FALSE}
# Mapping set up
# Load the packages
library(sf)
library(leaflet)
library(tidycensus)

census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")
install = TRUE
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) |>
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts |>
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"  # This is a fallback in case there are other values
  ))

head(nyc_tracts)

```

```{r, warning=FALSE}
nyc_tracts = nyc_tracts |>
  rename(geoid = GEOID)

redlining_sf = 
  nyc_tracts |>
  left_join(redlining_snap, by = "geoid")

```

## Redlining Exploration

Below is the map detailing the redlining grades given to areas in NYC. The corresponding colors and grades are in the legend. There are areas that our data set does not have information for, which are colored grey.

```{r redlining, echo=FALSE, warning=FALSE}

pal <- colorFactor(
  palette = c("#208e56",  
              "#e4e44c", 
              "#1595b4", 
              "#ef647b"),
  domain = redlining_sf$red_grade
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~pal(red_grade),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.9,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, "<br>Redlining Interval: ", interval2020, "<br>Redlining Score (2020): ", hrs2020)
  ) |>
  addLegend(
    pal = pal,
    values = ~red_grade,
    opacity = 0.7,
    title = "Redlining Category",
    position = "bottomright"
  )

```
This map effectively captures the historical spatial boundaries of redlining across NYC boroughs, showing how systemic inequities were distributed geographically.

- Manhattan and Brooklyn show stark contrasts between privileged (green/blue) and marginalized (yellow/red) areas.
- The Bronx and Brooklyn have the highest concentration of historically "hazardous" (D) zones, reflecting long-standing inequities.
- Queens and Staten Island had relatively fewer redlined areas, suggesting different historical patterns of urban development and disinvestment.

### Distribution of Redlining Scores in Boroughs

Lets look at Redlining Scores in a different way

```{r, fig.align='center'}
redlining = redlining |>
  mutate(
    borough = case_when(
      fip == 36061 ~ "Manhattan",
      fip == 36047 ~ "Brooklyn",
      fip == 36081 ~ "Queens",
      fip == 36005 ~ "Bronx",
      fip == 36085 ~ "Staten Island",
      TRUE ~ "Unknown"))

  ggplot(redlining, aes(x = borough, y = hrs2020, fill = borough)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  coord_flip() +
   stat_summary(
     fun = mean, geom = "point", color = "blue", size = 3, shape = 18) +
  stat_summary(
    fun = median, geom = "point", color = "purple", size = 3, shape = 16) +
  labs(
    title = "Distribution of Redlining Scores by NYC Borough",
    x = "Borough",
    y = "Redlining Score",
    caption = "Mean (Blue Diamond), Median (Purple Circle)") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() 

```

This graph shows different details. The distribution of the scores are fairly different for each of the boroughs, with Manhattan having the largest range and Queens having the smallest. Additionally, the mean and median redlining score for each of the borough are almost overlapping/the same. However, Manhattan has pretty different mean and median (mean closer to 3 and median closer to 4).

### Redlining Grade Count by Borough

```{r, fig.align='center'}
set2_palette = RColorBrewer::brewer.pal(n = 4, name = "Set2") 

redlining |> 
  count(borough, red_grade) |> 
  mutate(borough = fct_reorder(borough, n)) |> 
  plot_ly(
    x = ~borough,
    y = ~n,
    color = ~red_grade,
    type = "bar",
    colors = set2_palette
  ) |> 
  layout(
    title = "Count of Redlining Grades by Borough",
    xaxis = list(title = "Borough"),
    yaxis = list(title = "Count"),
    barmode = "stack" 
  )
```

This graph shows the grade count in each of the boroughs. Staten Island, Queens, and Bronx have the most C grade in their neighborhood (followed by D grade in second place). Manhattan and Brooklyn have grade D as the most frequent. This is reflective of what was previously observed in the violin plot. 

## SNAP Exploration and Demographics

### SNAP Enrollment

```{r, warning=FALSE}

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = redlining_sf$ph_snap,
  na.color = "grey"
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addPolygons(
    fillColor = ~snap_pal(ph_snap),
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste0("Census Tract: ", geoid, 
                    "<br>SNAP Enrollment (%): ", round(ph_snap, 2))
  ) |>

  addLegend(
    pal = snap_pal,
    values = ~ph_snap,
    opacity = 0.7,
    title = "SNAP Enrollment (%)",
    position = "bottomright"
  ) 
```
This map illustrates the strong correlation between historical redlining and present-day SNAP enrollment, underscoring how systemic inequities continue to influence food assistance needs across NYC.

- The Bronx stands out with the highest overall SNAP enrollment, aligning with its high concentration of historically redlined zones and lower-income households.
- Brooklyn also shows significant SNAP participation in neighborhoods historically marked as “hazardous.”
- Queens and Staten Island have lower overall SNAP participation, which reflects fewer redlined areas and more suburban development patterns.


### SNAP Total % of Households within Redline Grade

```{r, fig.align='center'}
redlining_snap |>
  separate(
    name,
    into = c("census_tract", "county", "state"),
    sep = "; ",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(
    borough = case_when(
      county == "Queens County" ~ "Queens",
      county == "Bronx County" ~ "Bronx",
      county == "New York County" ~ "Manhattan",
      county == "Kings County" ~ "Brooklyn",
      county == "Richmond County" ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  ) |>
  group_by(red_grade) |>
  summarize(
    avg_ph_snap = mean(ph_snap, na.rm = TRUE)
  ) |>
  ungroup() |>
  plot_ly(
    x = ~red_grade,
    y = ~avg_ph_snap,
    type = "bar",
    color = ~red_grade,
    colors = set2_palette
  ) |>
  layout(
    title = "Average SNAP Percentage of Households by Redline Grade in 2022",
    xaxis = list(title = "Redline Grade"),
    yaxis = list(title = "Average SNAP Percentage"),
    showlegend = FALSE
  )
  
```

This graph depicts the average % of households on SNAP within the areas with the grades. Unsurprisingly, areas with redlining grade of D have the highest average % of households on SNAP (23.7%), followed by C (18.6%), then B (17.2) and A (5.7) with the lowest. 

### Race VS Redlinging Grade

This is a stacked bar chart showing the race distribution of households by historical redlining grades (A, B, C, D, NA) in New York City.

```{r Race, warning=FALSE}
redlining_long <- redlining_sf |>
  select(red_grade, ph_white, ph_black, ph_asian, ph_other_race) |>
  group_by(red_grade)|>
  summarise(pha_white = mean(ph_white, na.rm = TRUE),
            pha_black = mean(ph_black, na.rm = TRUE),
            pha_asian = mean(ph_asian, na.rm = TRUE),
            pha_other_race = mean(ph_other_race, na.rm = TRUE))|>
  pivot_longer(
    cols = c(pha_white, pha_black, pha_asian, pha_other_race),
    names_to = "race_group",
    values_to = "percent_household"
  ) |>
  mutate(race_group = case_when(
    race_group == "pha_white" ~ "White",
    race_group == "pha_black" ~ "Black",
    race_group == "pha_asian" ~ "Asian",
    race_group == "pha_other_race" ~ "Other"
  ))

ggplot(redlining_long, aes(x = red_grade, y = percent_household, fill = race_group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_brewer(palette = "Set2") + 
  theme_minimal() +
  labs(
    title = "Race Distribution (White, Black, Asian, Other) by Redlining Grade",
    x = "Redlining Grade",
    y = "Percentage of Households",
    fill = "Race Group"
  ) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

* Grade D ("Hazardous") areas have a higher concentration of Black and minority households compared to Grade A or B areas, which show a higher proportion of White households.
* Grade C and D areas reflect the legacy of racial segregation, where redlined zones were predominantly occupied by communities of color.

## Healthy Grocery Store Exploration

### Grocery Stores (Manhattan, Brooklyn and Bronx)

```{r, warning=FALSE}

snap_pal <- colorNumeric(
  palette = "YlGnBu",
  domain = redlining_sf$ph_snap,
  na.color = "grey"
)

leaflet(redlining_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircleMarkers(
    data = nyc_healthy_store,
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,  
    color = "#0f6921", 
    stroke = FALSE,
    fillOpacity = 0.7,
    label = ~paste0("Store Name: ", store_name,
                    "<br>Borough: ", borough,
                    "<br>ZIP Code: ", zip_code),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "auto"
    )
  ) |>

  addLegend(
    position = "bottomleft",
    colors = "#0f6921",
    labels = "Healthy Grocery Store",
    title = "Healthy Grocery Stores in Bronx",
    opacity = 0.7
  )

```
This map provide a focused exploration of healthy grocery stores in three boroughs: Manhattan, Brooklyn, and the Bronx.
- The density of healthy grocery stores in Bronx is uneven across neighborhoods, with higher concentrations in specific areas.
- In Brooklyn, healthy grocery stores are relatively well-distributed, but there are clear clusters in central and northern neighborhoods. Southern Brooklyn appears less represented in this dataset, potentially reflecting disparities in access.


## Health Outcomes Exploration

### Health Outcomes in NYC

```{r, warning=FALSE, message=FALSE}
health_data =
  place_cleaned |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()

health_data_long <- health_data %>%
  drop_na(red_grade, obesity, diabetes, highchol) %>%
  pivot_longer(cols = c(obesity, diabetes, highchol), 
               names_to = "condition", values_to = "rate")

census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")  # API key for Census API
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) %>%
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts %>%
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"
  )) |>
  janitor::clean_names()

health_spatial_data <- nyc_tracts %>%
  left_join(health_data_long, by = "geoid")  # Make sure the 'geoid' column matches in both datasets

leaflet(health_spatial_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%  
  addPolygons(
    fillColor = ~colorNumeric(palette = "YlOrRd", domain = health_spatial_data$rate)(rate),  # Color based on 'rate'
    weight = 1,
    opacity = 1,
    color = "black",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    # Modify the label to show 'condition' and 'rate'
    label = ~paste0("Condition: ", condition,
                    "<br>Rate: ", round(rate, 2))
  ) %>%
  addLegend(
    pal = colorNumeric(palette = "YlOrRd", domain = health_spatial_data$rate),  # Legend for color scale
    values = ~rate,
    title = "Health Condition Rate",
    position = "bottomright"
  )
```

### Health Outcome vs Redlining Scores

```{r, message=FALSE, fig.align='center'}
health_data =
  place_cleaned |>
  mutate(geoid = as.character(geoid)) |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()

health_data |>
  drop_na(red_grade, obesity, diabetes, highchol) |>
  select(red_grade, obesity, diabetes, highchol) |>
  group_by(red_grade) |>
  mutate(
    obesity_rate = mean(obesity, na.rm = FALSE),
    diabetes_rate = mean(diabetes, na.rm = FALSE),
    highcholesterol_rate = mean(highchol, na.rm = FALSE)
  ) |>
  select(red_grade, obesity_rate, diabetes_rate, highcholesterol_rate) |>
  pivot_longer(cols = c(obesity_rate, diabetes_rate, highcholesterol_rate), 
               names_to = "condition", values_to = "rate") |>
  ggplot(aes(x = condition, y = rate, fill = condition)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  facet_wrap(~red_grade) + 
  labs(
    title = "Diabetes, Obesity, and High Cholesterol Rates by Redlining Grade",
    x = "Condition",
    y = "Rate"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

```

The graph shows how redlining grades correlate with health outcomes (diabetes, obesity, and high cholesterol) in New York City. The neighborhoods with A grades have slightly lower rates of diabetes and obesity, but D areas suffer from higher rates of these conditions. High cholesterol rates are interestingly higher in better-graded areas, suggesting that factors such as access to healthcare and preventive services may play a significant role in these health outcomes.

# Statistical Analysis

We import the data set and merge the snap enrollment, redlining and place data together to create the dataset `final analysis`. We also created a variable to represent the percentage of household with minority population (1-percentage of household with white race) as `minority`. 

```{r message=FALSE, warning=FALSE}
final_analysis = 
  redlining_snap|>
  left_join(place_cleaned, by = "geoid")|>
  filter(!is.na(obesity))|>
  mutate(
    minority = 100-ph_white
  )
```


## Desciptive Characteristics Based on Redlining Grades

### Total count of census tract for each HOLC grade

```{r}
final_analysis |>
  group_by(red_grade)|>
  summarise(count = n())|>
   knitr::kable()
```

The number of A-grade census tract is the smallest (n=36) while the number of C-grade census tract is the biggest(n=1003). 

### Comparison of HOLC grades 

```{r}
redlining_grades=
final_analysis |>
  group_by(red_grade) |>
  summarise(
    avg_snap = mean(ph_snap, na.rm = TRUE),
    sd_snap = sd(ph_snap, na.rm = TRUE),
    avg_elderly = mean(ph_elderly,na.rm=TRUE),
    sd_elderly = sd(ph_elderly, na.rm = TRUE),
    avg_child = mean(ph_with_children, na.rm = TRUE),
    sd_child = sd(ph_with_children, na.rm = TRUE),
    avg_poverty = mean(ph_below_poverty, na.rm = TRUE),
    sd_poverty = sd(ph_below_poverty, na.rm = TRUE),
    avg_disability = mean(ph_disability, na.rm = TRUE),
    sd_disability = sd(ph_disability, na.rm = TRUE),
    avg_white = mean(ph_white, na.rm = TRUE),
    sd_white = sd(ph_white, na.rm = TRUE),
    avg_black = mean(ph_black, na.rm = TRUE),
    sd_black = sd(ph_black, na.rm = TRUE),
    avg_asian = mean(ph_asian, na.rm = TRUE),
    sd_asian = sd(ph_asian, na.rm = TRUE),
    avg_otherrace = mean(ph_other_race, na.rm = TRUE),
    sd_otherrace = sd(ph_other_race, na.rm = TRUE),
    avg_hispanic = mean(ph_hispanic, na.rm = TRUE),
    sd_hispanic = sd(ph_hispanic, na.rm = TRUE),
    avg_nowork = mean(ph_no_work, na.rm = TRUE),
    sd_nowork = sd(ph_no_work, na.rm = TRUE),
    avg_1work = mean(ph_1_work, na.rm = TRUE),
    sd_1work = sd(ph_1_work, na.rm = TRUE),
    avg_2work = mean(ph_2_work, na.rm = TRUE),
    sd_2work = sd(ph_2_work, na.rm = TRUE),
    avg_obesity = mean(obesity, na.rm = TRUE),
    sd_obesity = sd(obesity, na.rm = TRUE),
    avg_highchol = mean(highchol, na.rm = TRUE),
    sd_highchol = sd(highchol, na.rm = TRUE),
    avg_diabetes = mean(diabetes, na.rm = TRUE),
    sd_diabetes = sd(diabetes, na.rm = TRUE)
  )|> pivot_longer(
    cols = -red_grade,
    names_to = c("stat", "variable"),
    names_sep = "_"
  )|>
  pivot_wider(
    names_from = c(red_grade, stat), 
    values_from = value,              
    names_glue = "{red_grade}_{stat}"
  )

anova = function(dataset, outcomes, predictor) {
  results = list() 
  for (outcome in outcomes) {
    formula = as.formula(paste(outcome, "~", factor(predictor)))
    model= aov(
      formula,
      data = dataset
    )
    results[[outcome]] = broom::tidy(model)
  }
  
  combined_results = bind_rows(results, .id = "outcome")
  return(combined_results)
}

outcomes = c("ph_snap","ph_with_children", "ph_elderly","ph_below_poverty", "ph_disability", "ph_white", "ph_black", "ph_asian", "ph_other_race", "ph_hispanic", "ph_no_work","ph_1_work", "ph_2_work", "obesity", "highchol", "diabetes")

baseline_char = 
anova(final_analysis, outcomes, "red_grade")|>
  filter(term == "red_grade")|>
  rename(variable = outcome)|>
  mutate(variable = case_when(
    variable == "ph_snap" ~ "snap",
    variable == "ph_with_children" ~ "child",
    variable == "ph_elderly" ~ "elderly",
    variable == "ph_below_poverty" ~ "poverty",
    variable == "ph_no_work" ~ "nowork",
    variable == "ph_white" ~ "white",
    variable == "ph_black" ~ "black",
    variable == "ph_asian" ~ "asian",
    variable == "ph_other_race" ~ "otherrace",
    variable == "ph_disability" ~ "disability",
    variable == "ph_hispanic" ~ "hispanic",
    TRUE ~ variable
  ))|>
  inner_join(redlining_grades, by="variable")|>
  select(variable, A_avg:D_sd,  p.value)


baseline_char|>
  filter(!variable %in% c("obesity","diabetes","highchol"))|>
  knitr::kable(digits = 3)
```

We computed the average and standard deviation of the different socio-demographic factors across the redlining grade groups. ANOVA was utilized to compare the mean across different census tract grades (A, B, C and D). The percentage of household enrolled in snap/ food stamp program at the census tract level was significantly different among groups with different redlining grades at 5% significance level. A-grade census tracts had the lowest average percentage of SNAP enrollment (5.72%, sd: 6.95%) and the average percentage of SNAP enrollment is the highest in D-grade census tracts (23.70, sd: 17.85). In addition, the groups with different redlining grades reported significant different for the average percentage regarding households with children smaller than 18 years, households with elderly grater than 60 years, households below the poverty line, households with people with disability, race and ethnicity, and employment status. 


### Comparing census tracts with HRS for Grade A with Grade D

```{r warning=FALSE, message=FALSE}
q1_vs_q4 = 
final_analysis|>
  filter(red_grade %in%c("A","D"))|>
  mutate()

q1q4_desc = 
q1_vs_q4|>
  group_by(red_grade)|>
  summarise(
    ph_snap = mean(ph_snap),
    ph_with_children = mean(ph_with_children),
    ph_elderly = mean(ph_elderly),
    ph_below_poverty = mean(ph_below_poverty),
    ph_disability = mean(ph_disability),
    minority = mean(minority),
    ph_no_work = mean(ph_no_work),
    obesity = mean(obesity),
    highchol = mean(highchol),
    diabetes = mean(diabetes)
  )|>
 pivot_longer(cols = -red_grade, names_to = "variable", values_to = "value")|>
 pivot_wider(names_from = red_grade, names_prefix = "grade_", values_from = value)|>
 mutate(
    mean_diff = grade_D - grade_A,
  )|>
  rename(outcome = variable)
```


```{r}
uni_reg = function(dataset, outcomes, predictor) {
  results = list() 
  for (outcome in outcomes) {
    formula = as.formula(paste(outcome, "~", predictor))
    model= t.test(
      formula,
      data = dataset
    )
    results[[outcome]] = broom::tidy(model)
  }
  
  combined_results = bind_rows(results, .id = "outcome")
  return(combined_results)
}

outcomes = c("ph_snap","ph_with_children", "ph_elderly","ph_below_poverty", "ph_disability", "ph_white", "ph_black", "ph_asian", "ph_no_work", "obesity", "highchol", "diabetes")


uni_reg(q1_vs_q4, outcomes, "red_grade")|>
  inner_join(q1q4_desc, by="outcome")|>
  select(outcome, grade_A, grade_D, mean_diff, p.value)|>
  knitr::kable(digits = 3)
```

Following the ANOVA test results, we wondered whether there is any significant difference across the least-redlined (A-grade) and most-redlined (D-grade) census tracts regarding snap enrollment, socio-demographic factors, and health outcomes. Based on the t-test outcome, D-grade census tracts had significantly higher average percentage of households enrolled for SNAP, average percentage of households with children, average percentage of households below the poverty line, average percentage of households with peoplw with diability, percentage of household with minority race compared to A-grade census tracts. Additionally, A-grade census tract had significantly higher average percentage of households with elderly, average percentage of households with no worker.

Regarding the health outcomes, the average prevalence of obesity is 7.97% higher, and the average prevalence for diabetes is 2.34% higher in the D-grade census tracts compared to A-grade census tracts. The average prevalence of high cholesterol level is 4.99% higher in the A-grade census tracts compared to D-grade census tracts. All of the difference in the average prevalence is statistically signficant. 


## Regression Analysis for Redlining and Food Insecurity

### Univariate Modelling

Based on the results for the baseline characteristics analysis, we computed a linear regression model to explore the association between redlining grades and precentage of households enrolled in SNAP. 

```{r}
# univariate analysis
lm(ph_snap~red_grade,final_analysis)|>
  broom::tidy()|>
  knitr::kable(digits= 3)
```

Based on the univariate model, we found the percentage of household enrolled in SNAP was 17.97%, 12.90% and 11.53% higher respectively in the D-grade, C-grade and B-grade census tracts compared to A-grade census tracts (all p-value <0.001). 

### Multivariate Modelling

```{r}
## multivariate model 1
red_snap_m1 = lm(ph_snap~red_grade+minority+ph_elderly+ph_disability+ph_no_work,final_analysis)
tidy_model = broom::tidy(red_snap_m1)
ci = confint(red_snap_m1)

tidy_model|>
  mutate(
    CI_Lower = ci[, 1],
    CI_Upper = ci[, 2],
  )|>
  select(term, estimate, p.value, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 1 Estimates with 95% Confidence Intervals")
```

When the association between redlining grades and SNAP enrollment evaluated using a linear regression model, D-grade census tracts had a 4.04% higher SNAP enrollment compared to A-grade census tracts (p = 0.037), indicating a statistically significant relationship. However, for B-grade and C-grade tracts, the differences were not statistically significant (p = 0.087 and p = 0.283 respectively). Among the covariates, a 1% increase in minority population was associated with a 0.16% increase in SNAP enrollment (p < 0.001). Additionally, the percentage of households with elderly members was negatively associated with SNAP enrollment (-0.33%, p < 0.001), while households with disabilities (+0.75%, p < 0.001) and those with no workers (+0.32%, p < 0.001) showed significant positive associations.

```{r}
## histogram for snap enrollment
ggplot(final_analysis, aes(x = ph_snap)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
    title = "Histogram of ph_snap",
    x = "SNAP Enrollment (ph_snap)",
    y = "Frequency"
  ) +
  theme_minimal()
```

The histogram shows that SNAP enrollment is right-skewed, with the majority of census tracts having lower percentages of households enrolled. Therefore, transformations were applied.

```{r}
## multivariate model 2
red_snap_m2 =
  final_analysis|>
  lm(log(ph_snap+1)~red_grade+minority+ph_elderly+ph_disability+ph_no_work, data=_)

tidy_model = broom::tidy(red_snap_m2)
ci = confint(red_snap_m2)

tidy_model|>
  mutate(
    Back_Transformed_Estimate = exp(estimate),
    CI_Lower = exp(ci[, 1]),
    CI_Upper = exp(ci[, 2]),
    p.value = ifelse(p.value<0.001, "<0.001", p.value)
  )|>
  select(term, estimate, p.value, Back_Transformed_Estimate, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 2 Coefficients with Back-Transformed Estimates and 95% Confidence Intervals")
```

We applied a log transformation to SNAP enrollment to better address skewness in the data, with results interpreted as multiplicative changes. Compared to A-grade tracts, D-grade tracts exhibited 1.82 times higher SNAP enrollment (p < 0.001). Similarly, B-grade and C-grade tracts showed increases of 1.78 times (p < 0.001) and 1.75 times (p < 0.001), respectively. Covariates such as minority population and households with disabilities remained significant. A 1% increase in minority population was associated with a 1.2% increase in SNAP enrollment (p < 0.001), while households with elderly members were associated with a 1.4% decrease in SNAP enrollment (p < 0.001).

```{r}
##multivariate model 3
red_snap_m3 = lm(log(ph_snap+1)~red_grade+ph_white+ph_black+ph_asian+ph_elderly+ph_disability+ph_no_work,final_analysis)

tidy_model = broom::tidy(red_snap_m3)
ci = confint(red_snap_m3)

tidy_model|>
  mutate(
    Back_Transformed_Estimate = exp(estimate),
    CI_Lower = exp(ci[, 1]),
    CI_Upper = exp(ci[, 2]),
    p.value = ifelse(p.value<0.001, "<0.001", p.value)
  )|>
  select(term, estimate, p.value, Back_Transformed_Estimate, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 3 Coefficients with Back-Transformed Estimates and 95% Confidence Intervals")
```

Then we included specific ethnic proportions (White, Black, and Asian) as covariates, refining the analysis. In this model, D-grade tracts had 1.94 times higher SNAP enrollment compared to A-grade tracts (p < 0.001). B-grade and C-grade tracts had increases of 1.85 times (p < 0.001) and 1.80 times (p < 0.001), respectively. Among ethnic covariates, a higher proportion of White households was associated with slightly lower SNAP enrollment, with a 1% increase in White population reducing enrollment by 1.9% (p < 0.001). Similarly, higher proportions of Black and Asian households also exhibited small negative associations. As in previous models, households with disabilities (+3.2%, p < 0.001) and no workers (+0.9%, p < 0.001) remained significant predictors of higher SNAP enrollment. 


### Cross Validation and Model Selection

To assess the predictive performance of the three multivariate models, Monte Carlo cross-validation was conducted, calculating the Root Mean Square Error (RMSE) for each model across multiple training and testing splits.


```{r}
library(modelr)

cv_df =
  crossv_mc(final_analysis, 100)

cv_df =
cv_df|>
  mutate(model_1 = map(train, \(df)  
                       lm(ph_snap~red_grade+minority+ph_elderly+ph_disability+ph_no_work,
                          data=df)),
         model_2 = map(train, \(df) 
                       lm(log(ph_snap+1)~red_grade+minority+ph_elderly+ph_disability
                          +ph_no_work,data=df)),
         model_3 = map(train, \(df) 
                       lm(log(ph_snap+1)~red_grade+ph_white+ph_black+ph_asian+ph_elderly
                          +ph_disability+ph_no_work, data=df))
           )|>
  mutate(
    rmse_m1 = map2_dbl(model_1, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_m2 = map2_dbl(model_2, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_m3 = map2_dbl(model_3, test, \(mod, df) rmse(model = mod, data = df))
  )

cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()+
  labs(title = "Model 1 vs. Model 2 & 3")+
  theme_minimal()
```

Model 1, which used a simple linear approach without log transformation, demonstrated the highest RMSE values, averaging over 10 units. This poor performance indicates that the linearity assumption of Model 1 poorly captures the variability in SNAP enrollment.

Conversely, Model 2 and Model 3, both of which log-transformed SNAP enrollment to account for its right-skewed distribution, exhibited significantly lower RMSE values. This suggests that these models provided a much better fit to the data and were more reliable in predicting SNAP enrollment.

```{r}
cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  filter(model == "m2" | model== "m3")|>
  ggplot(aes(x = model, y = rmse)) + geom_violin()+
  labs(title = "Model 2 vs. Model 3")+
  theme_minimal()
```

Focusing on the violin plot comparing Model 2 and Model 3, the RMSE values for both models are clustered closely together, with Model 3 showing slightly lower RMSE on average. Model 3, which incorporates detailed racial and ethnic composition variables, performed marginally better than Model 2, which used a simplified “minority” variable to summarize racial and ethnic factors. This improvement indicates that accounting for specific racial variations enhances the model’s ability to explain and predict SNAP enrollment.

Therefore, model 3 provides the most robust framework for understanding the relationship between redlining, demographic factors, and food insecurity.


## Association of Health Outcomes and Redlining

### Preview the Health Outcomes Data
```{r}
baseline_char|>
  filter(variable %in% c("obesity","diabetes","highchol"))|>
  mutate(p.value = ifelse(p.value<0.001, "<0.001", p.value)) |> 
  knitr::kable(digits = 3)

final_analysis |>
  select(red_grade, obesity, diabetes, highchol) |>
  pivot_longer(cols = c(obesity, diabetes, highchol), 
               names_to = "outcome", 
               values_to = "value") |> 
  ggplot(aes(x = red_grade, y = value, fill = outcome)) +
  geom_boxplot() +
  facet_wrap(~ outcome, scales = "free_y") +
  labs(title = "Boxplot of Health Outcomes by Redlining Grade",
       x = "Redlining Grade",
       y = "Outcome Value")
```

- Obesity prevalence increases with worsening redlining grades, from 20.74% in A-grade tracts to 28.71% in D-grade tracts (p < 0.001). C-grade and B-grade tracts show averages of 27.51% and 25.92%, respectively. The boxplot reveals higher median values and greater variability in obesity for more redlined areas.

- High cholesterol prevalence decreases with worsening redlining grades, from 37.28% in A-grade tracts to 32.29% in D-grade tracts (p < 0.001). B-grade and C-grade tracts have averages of 34.50% and 34.14%, with tighter distributions in less-redlined areas.

- Diabetes prevalence also rises with worsening redlining grades, increasing from 9.25% in A-grade tracts to 12.36% in C-grade tracts and 11.58% in D-grade tracts (p < 0.001). B-grade tracts fall in between at 11.11%, with greater variability observed in more redlined areas.


### Univariate Modelling

```{r}
diabetes_model = lm(diabetes ~ red_grade, data = final_analysis)
diabetes_model |> 
  broom::tidy()|> 
  knitr::kable(digits = 3, caption = "Univariate Regression Model for Diabetes")
```

In the univariable diabetes model, the results indicate a similar trend, with redlining grades playing a significant role. The intercept for grade A was 9.247 (p < 0.001), while the estimates for grade B, grade C, and grade D were 1.866, 3.108, and 2.337, respectively. While the relationship is positive, it is less pronounced than for obesity. The coefficients show a gradual increase in diabetes levels as the redlining grade worsens, and the p-values for all estimates were statistically significant (p < 0.05), further solidifying the link between redlining and diabetes.

```{r}
chol_model = lm(highchol ~ red_grade, data = final_analysis) 
chol_model |> 
  broom::tidy()|> 
  mutate(p.value = ifelse(p.value<0.001, "<0.001", p.value)) |> 
  knitr::kable(digits = 3, caption = "Univariate Regression Model for High Cholesterol")
```

The results from the univariable high cholesterol model showed a clear relationship between redlining grades and high cholesterol. Specifically, individuals from areas with worse redlining grades (e.g., grades C and D) are more likely to have higher high cholesterol outcomes. The intercept of 37.275 (p < 0.001) indicates a baseline level of high cholesterol for individuals in grade A areas. However, when moving to B, C, and D, the estimated effects were negative, with estimates of -2.775, -3.133, and -4.989, respectively, all with p-values less than 0.001. This trend suggests that, contrary to what might be expected, individuals from higher redlining grades may have lower high cholesterol scores, potentially due to systemic barriers and less access to health resources that prevent the management of high cholesterol.

```{r}
obesity_model = lm(obesity ~ red_grade, data = final_analysis) 
obesity_model|> 
  broom::tidy()|> 
  mutate(p.value = ifelse(p.value<0.001, "<0.001", p.value)) |> 
  knitr::kable(digits = 3, caption = "Univariate Regresion Model for Obesity")
```

In the univariable obesity model, the relationship between redlining and obesity outcomes is more directly significant. The intercept of 20.736 (p < 0.001) provides the baseline obesity rate for individuals in grade A areas. For grade B, grade C, and grade D, the estimates were 5.187, 6.770, and 7.974, respectively (all with p < 0.001). These results showed that as the redlining grade worsens, obesity prevalence increases. This finding aligns with previous research suggesting that neighborhoods with poorer infrastructure, limited access to healthy food, and reduced access to healthcare services may exacerbate obesity rates. The statistically significant p-values for all grades further support that redlining has a substantial effect on obesity outcomes.


Thus, we extended analysis and performed multivariable regression models that adjusted for additional covariates, including race (ph_white, ph_black, ph_asian), age (ph_elderly), disability (ph_disability), and employment status (ph_no_work). The goal of these models was to assess whether the relationships observed between redlining grades and health outcomes would hold after accounting for these important socio-demographic factors.


### Multivariate Modelling

```{r}
diabetes_adj_model = lm(log(diabetes)~red_grade+ph_white+ph_black+ph_asian+ph_elderly+ph_disability+ph_no_work,final_analysis)
diabetes_adj_model |> 
  broom::tidy()|> 
  knitr::kable(digits = 3, caption = "Multivariate Regression Model for Diabetes")
```

The adjusted model for diabetes revealed no significant differences in diabetes prevalence between redlining grades after accounting for covariates such as racial composition, elderly population, disability, and no-work households. The estimates for B-grade (0.022, p = 0.520), C-grade (0.057, p = 0.092), and D-grade (-0.008, p = 0.825) were not statistically significant. Among the covariates, the proportion of White, Black, and Asian populations were significantly associated with diabetes prevalence. Specifically, a 1% increase in the White population was associated with a 0.9% decrease in diabetes prevalence (p < 0.001), while increases in Black and Asian populations were also negatively associated but to a lesser extent. Other significant covariates included the proportion of elderly households (+0.6%, p < 0.001), disability (+0.7%, p < 0.001), and no-work households (+0.5%, p < 0.001), all of which were positively associated with diabetes prevalence.

```{r}
highcol_adj_model = lm(log(highchol)~red_grade+ph_white+ph_black+ph_asian+ph_elderly+ph_disability+ph_no_work,final_analysis)
highcol_adj_model |> 
  broom::tidy()|> 
  knitr::kable(digits = 3, caption = "Multivariate Regression Model for High Cholesterol")
```

The adjusted model for high cholesterol showed a significant decrease in prevalence in B-grade (-0.034, p = 0.015), C-grade (-0.047, p = 0.001), and D-grade (-0.059, p < 0.001) tracts compared to A-grade tracts. Among the covariates, the proportion of Asian households was positively associated with high cholesterol prevalence (+0.2%, p < 0.001), while the proportion of elderly households was strongly positively associated (+2.5%, p < 0.001). Disability (+0.1%, p < 0.001) and no-work households (+0.2%, p < 0.001) also showed positive associations. Interestingly, the proportions of White and Black populations did not exhibit significant associations with high cholesterol prevalence.

```{r}
obesity_adj_model = lm(log(obesity)~red_grade+ph_white+ph_black+ph_asian+ph_elderly+ph_disability+ph_no_work,final_analysis)
obesity_adj_model |> 
  broom::tidy()|> 
  knitr::kable(digits = 3, caption = "Multivariate Regression Model for Obesity")
```

The adjusted model for obesity indicated significant differences across redlining grades, with obesity prevalence higher in B-grade (0.101, p < 0.001), C-grade (0.151, p < 0.001), and D-grade (0.118, p < 0.001) tracts compared to A-grade tracts. Among the covariates, the proportion of White households was negatively associated with obesity prevalence, with a 0.5% decrease per 1% increase in the White population (p < 0.001). Conversely, increases in Black and Asian populations were also negatively associated, with estimates of -0.1% (p < 0.001) and -1.0% (p < 0.001), respectively. Other significant covariates included disability (+0.5%, p < 0.001) and no-work households (+0.1%, p = 0.011), which were positively associated with obesity prevalence. The proportion of elderly households was negatively associated (-0.1%, p = 0.001).


# Discussion

Our analysis demonstrates a significant association between historical redlining and food insecurity in New York City, highlighting how discriminatory housing policies have shaped access to resources and health outcomes at the census tract level. Neighborhoods historically graded as “declining” or “hazardous” were more likely to experience limited access to adequate food compared to those classified as “desirable” by HOLC. These findings underscore the enduring legacy of systemic disinvestment in marginalized communities.

Redlining has entrenched structural inequalities, disproportionately affecting racial minorities and individuals with low socioeconomic status. Redlined neighborhoods have higher proportions of racial minorities, households with people with disabilities, and families living under the poverty line. In terms of health outcomes, redlined neighborhoods exhibit significantly higher prevalence rates of obesity but a surprisingly lower prevalence of high cholesterol compared to more privileged neighborhoods. Diabetes, however, showed no significant direct association with redlining in adjusted models. These results suggest a complex interplay of historical policies, food access, and health outcomes that requires further investigation.


## Implications for Public Health and Policy

These findings underscore the importance of addressing historical and systemic inequities through targeted public health interventions and policy reforms. The concentration of food insecurity and health disparities in redlined neighborhoods calls for investment in community-based programs aimed at increasing access to affordable, nutritious food. Expanding the availability of healthy grocery stores, improving public transportation, and incentivizing local food production could mitigate the challenges faced by these communities.

Additionally, policies should prioritize improving healthcare access and preventive services in historically redlined neighborhoods. Initiatives such as mobile health clinics, community health education, and subsidies for preventive care can help address the health disparities observed in these areas. Policymakers must also recognize the role of urban planning in shaping social determinants of health. Collaborative efforts involving public health agencies, city planners, and community organizations are essential to dismantling the structural barriers created by redlining and fostering equitable urban development.

## Strength and Limitations

One strength of this study is the integration of multiple data sources, allowing for a comprehensive analysis of the socioeconomic and health dimensions of food insecurity. By combining visualizations and statistical models, this research provides insights into the legacy of redlining and its implications for modern-day disparities.

While this study provides valuable insights, it is not without limitations.The right-skewed distribution of SNAP enrollment posed challenges to modeling assumptions, which we addressed through log transformations. Additionally, the cross-sectional design limits causal interpretations, and aggregated census tract-level data may obscure individual variations. The cross-sectional nature of the analysis limits the ability to infer causality, and the use of aggregated census tract-level data may mask individual-level variations. Additionally, the models assume that historical redlining is uniformly experienced across neighborhoods, which may oversimplify the nuances of community dynamics and contemporary urban policies. Future research should explore longitudinal data to examine how neighborhood changes over time mediate the relationship between historical redlining and food insecurity.


# Conclusion

Our findings reveal the persistent and multifaceted impact of historical redlining on food insecurity and health disparities in New York City. Addressing these inequities requires systemic changes that go beyond individual interventions, focusing on dismantling structural barriers and promoting community resilience. By prioritizing equity in urban development and public health policies, it is possible to create sustainable, healthier environments for all New Yorkers, particularly those in historically marginalized communities.

