---
title: "Final Report"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

## Motivation

The practice of redlining, where certain neighborhoods were systematically denied access to resources based on racial and economic factors, has had long-lasting effects on urban communities. In New York City, the legacy of redlining continues to shape access to vital services such as healthy food and public assistance programs. The motivation for this project is to explore how these historical practices have impacted modern-day disparities in access to healthy grocery stores and enrollment in the Supplemental Nutrition Assistance Program (SNAP).

Our primary goal is to investigate the relationship between the neighborhoods that were historically redlined and their current access to essential resources. We seek to answer questions about the extent to which redlining has contributed to food deserts, areas where healthy food options are scarce, and whether these communities have disproportionately higher rates of SNAP enrollment. By understanding these connections, we aim to uncover the ways in which systemic inequalities persist in shaping the health and well-being of New Yorkers, and to contribute to ongoing discussions about potential policy solutions to address these disparities.

This project is driven by a desire to shed light on the intersection of urban planning, socioeconomic disparities, and public health, and to highlight the role of policy in creating both opportunities and barriers for communities that continue to experience the effects of redlining. Through this exploration, we hope to contribute valuable insights to the broader conversation about equity in urban environments and the long-term impacts of discriminatory practices on marginalized populations.

## Initial Questions

- **Neighborhood:** What areas in NYC were most heavily impacted by redlining? How do the demograpics within these areas compare?
- **Grocery:** Are there fewer healthy grocery stores in neighborhoods that were historically redlined compared to other areas? How does access to health grocery stores correlate with health outcomes in these neighborhoods?
- **SNAP:** Is there a higher rate of SNAP enrollment in areas that were historically redlined? Do these areas face more challenges accessing SNAP benefits? 
- **Analysis:**
  - How does redlining classification (e.g., grade A, B, C, D) predict the density of healthy grocery stores in a neighborhood? What is the relationship between redlining and SNAP enrollment rates across neighborhoods in NYC?
  - Are there other factors (e.g., income levels) that mediate the relationship between redlining and food access?

## Data

### Data Source 

### Data Cleaning

### Data Merge

## Exploratory Analysis

### Visualizations:

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(plotly)
```

```{r, echo=FALSE}
redlining_snap =
  redlining_snap |>
  mutate(borough = case_when(
    str_detect(name, "Bronx") ~ "Bronx",
    str_detect(name, "Kings") ~ "Brooklyn",
    str_detect(name, "New York County") ~ "Manhattan",
    str_detect(name, "Queens") ~ "Queens",
    str_detect(name, "Richmond") ~ "Staten Island",
    TRUE ~ "Other"
  ))

health_data =
  place_cleaned |>
  mutate(geoid = as.character(geoid)) |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()
```

#### Distribution of Redlining Scores:

```{r}
redlining_snap |>
  plot_ly(y = ~hrs2020, color = ~borough, type = "box", colors = "viridis") |>
  layout(title = "Distribution of Redlining Scores in Each Borough",
         xaxis = list(title = 'Borough'),
         yaxis = list(title = 'Redlining Score'),
        showlegend = FALSE)
```

#### Obesity vs Redlining Score:

```{r, echo=FALSE}
health_data =
  health_data |>
  filter(!is.na(obesity) & !is.na(hrs2020))

health_data |>
  plot_ly(x = ~hrs2020,y = ~obesity, type = 'scatter', mode = 'markers', 
             color = ~borough,
             hoverinfo = 'text', 
             text = ~paste('Borough:', borough, '<br>Obesity Rate:', obesity, '%', '<br>Redlining Score:', hrs2020)) |>
  layout(
    title = "Obesity Rate vs Redlining Score",
    xaxis = list(title = "Redlining Score (HRS 2020)"),
    yaxis = list(title = "Obesity Rate (%)"),
    showlegend = FALSE
  ) |>
  facet_wrap(~borough)

#This is another version
health_data |>
  ggplot(aes(y = obesity, x = hrs2020, color = borough)) +
  geom_point() +
  geom_smooth(color = "black") +
  facet_wrap(~ borough, scales = "free") +
  labs(
    title = "Obesity Rate vs Redlining Score",
    x = "Obesity Rate (%)",
    y = "Redlining Score (HRS 2020)"
  ) +
  theme_minimal()
```

The exploratory analysis of obesity rates against redlining scores reveals a consistent positive trend across most of New York City's boroughs, with areas of historical redlining showing higher obesity rates. While Manhattan presents a weaker relationship, other boroughs like the Bronx, Brooklyn, Queens, and Staten Island demonstrate stronger associations, suggesting that the legacy of redlining may contribute to health disparities. 

## Regression Analysis

## Discussion









