---
title: "Statistical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(modelr)

SNAP = 
  read_csv(file = "./Datasets/ACS_SNAP_ENROLLMENT.csv")|>
  janitor::clean_names()|>
  select(
    geo_id:s2201_c01_001e,
    s2201_c02_002e,
    s2201_c02_009e,
    s2201_c02_015e,
    s2201_c02_021e:s2201_c02_028e,
    s2201_c02_032e,
    s2201_c02_036e:s2201_c02_038e,
    s2201_c04_001e)|>
  select(-ends_with("m"))|>
  slice(-1)|>
  mutate(geoid = str_remove(geo_id, ".*US"))|>
  rename(
     total_ct_households = s2201_c01_001e,
     ph_elderly = s2201_c02_002e,
     ph_with_children = s2201_c02_009e,
     ph_no_children = s2201_c02_015e,
     ph_below_poverty = s2201_c02_021e,
     ph_above_poverty = s2201_c02_022e,
     ph_disability = s2201_c02_023e,
     ph_no_disability = s2201_c02_024e,
     ph_white = s2201_c02_025e,
     ph_black = s2201_c02_026e,
     ph_asian = s2201_c02_028e,
     ph_hispanic = s2201_c02_032e,
     ph_no_work = s2201_c02_036e,
     ph_1_work = s2201_c02_037e,
     ph_2_work = s2201_c02_038e,
     ph_snap = s2201_c04_001e
  )|>
  mutate(across(total_ct_households:ph_snap, as.numeric),
         ph_other_race = 100-ph_white-ph_black-ph_asian,
         ph_non_hispanic = 100-ph_hispanic)|>
  select(geoid,name,ph_snap,everything())|>
  select(-geo_id)

redlining = 
  read_excel("./Datasets/Historic Redlining Score 2020B.xlsx")|>
  janitor::clean_names()|>
  select(-cbsa)|>
  mutate(fip = substr(geoid20, 1, 5),
         interval2020 = as.character(interval2020),
         red_grade = case_when(
           hrs2020 < 1.5 ~ "A",
           hrs2020 < 2.5 ~ "B",
           hrs2020 < 3.5 ~ "C",
           TRUE ~ "D"
         ))|>
  filter(fip %in% c("36005", "36047", "36061", "36081", "36085"))|>
  rename(geoid = geoid20)

redlining_snap = 
  redlining|>
  left_join(SNAP, by = "geoid")|>
  filter(is.na(ph_snap)==FALSE)

place_crude =
  read_csv("./Datasets/PLACES.csv")|>
  janitor::clean_names()|>
  rename(geoid = location_id)|>
  select(geoid, measure:data_value, measure_id)


place_cleaned=
  place_crude|>
  filter(measure_id %in% c("DIABETES", "HIGHCHOL", "OBESITY"))|>
  pivot_wider(
    id_cols = geoid,
    names_from = measure_id,
    values_from = data_value
  )|>
  janitor::clean_names()|>
  mutate(geoid = as.character(geoid))

# match place data with redlining_snap 
final_analysis = 
  redlining_snap|>
  left_join(place_cleaned, by = "geoid")|>
  filter(!is.na(obesity))|>
  mutate(
    minority = 100-ph_white
  )
```


## Desciptive Characteristics Based on Redlining Grades

### Total count of census tract for each HOLC grade 

```{r}
final_analysis |>
  group_by(red_grade)|>
  summarise(count = n())|>
   knitr::kable()
```

### Comparison of HOLC grades 

```{r}
redlining_grades=
final_analysis |>
  group_by(red_grade) |>
  summarise(
    avg_snap = mean(ph_snap, na.rm = TRUE),
    sd_snap = sd(ph_snap, na.rm = TRUE),
    avg_elderly = mean(ph_elderly,na.rm=TRUE),
    sd_elderly = sd(ph_elderly, na.rm = TRUE),
    avg_child = mean(ph_with_children, na.rm = TRUE),
    sd_child = sd(ph_with_children, na.rm = TRUE),
    avg_poverty = mean(ph_below_poverty, na.rm = TRUE),
    sd_poverty = sd(ph_below_poverty, na.rm = TRUE),
    avg_disability = mean(ph_disability, na.rm = TRUE),
    sd_disability = sd(ph_disability, na.rm = TRUE),
    avg_white = mean(ph_white, na.rm = TRUE),
    sd_white = sd(ph_white, na.rm = TRUE),
    avg_black = mean(ph_black, na.rm = TRUE),
    sd_black = sd(ph_black, na.rm = TRUE),
    avg_asian = mean(ph_asian, na.rm = TRUE),
    sd_asian = sd(ph_asian, na.rm = TRUE),
    avg_otherrace = mean(ph_other_race, na.rm = TRUE),
    sd_otherrace = sd(ph_other_race, na.rm = TRUE),
    avg_hispanic = mean(ph_hispanic, na.rm = TRUE),
    sd_hispanic = sd(ph_hispanic, na.rm = TRUE),
    avg_nowork = mean(ph_no_work, na.rm = TRUE),
    sd_nowork = sd(ph_no_work, na.rm = TRUE),
    avg_1work = mean(ph_1_work, na.rm = TRUE),
    sd_1work = sd(ph_1_work, na.rm = TRUE),
    avg_2work = mean(ph_2_work, na.rm = TRUE),
    sd_2work = sd(ph_2_work, na.rm = TRUE),
    avg_obesity = mean(obesity, na.rm = TRUE),
    sd_obesity = sd(obesity, na.rm = TRUE),
    avg_highchol = mean(highchol, na.rm = TRUE),
    sd_highchol = sd(highchol, na.rm = TRUE),
    avg_diabetes = mean(diabetes, na.rm = TRUE),
    sd_diabetes = sd(diabetes, na.rm = TRUE)
  )|> pivot_longer(
    cols = -red_grade,
    names_to = c("stat", "variable"),
    names_sep = "_"
  )|>
  pivot_wider(
    names_from = c(red_grade, stat), 
    values_from = value,              
    names_glue = "{red_grade}_{stat}"
  )

anova = function(dataset, outcomes, predictor) {
  results = list() 
  for (outcome in outcomes) {
    formula = as.formula(paste(outcome, "~", factor(predictor)))
    model= aov(
      formula,
      data = dataset
    )
    results[[outcome]] = broom::tidy(model)
  }
  
  combined_results = bind_rows(results, .id = "outcome")
  return(combined_results)
}

outcomes = c("ph_snap","ph_with_children", "ph_elderly","ph_below_poverty", "ph_disability", "ph_white", "ph_black", "ph_asian", "ph_other_race", "ph_hispanic", "ph_no_work","ph_1_work", "ph_2_work", "obesity", "highchol", "diabetes")

baseline_char = 
anova(final_analysis, outcomes, "red_grade")|>
  filter(term == "red_grade")|>
  rename(variable = outcome)|>
  mutate(variable = case_when(
    variable == "ph_snap" ~ "snap",
    variable == "ph_with_children" ~ "child",
    variable == "ph_elderly" ~ "elderly",
    variable == "ph_below_poverty" ~ "poverty",
    variable == "ph_no_work" ~ "nowork",
    variable == "ph_1_work" ~ "1work",
    variable == "ph_2_work" ~ "2work",
    variable == "ph_white" ~ "white",
    variable == "ph_black" ~ "black",
    variable == "ph_asian" ~ "asian",
    variable == "ph_other_race" ~ "otherrace",
    variable == "ph_disability" ~ "disability",
    variable == "ph_hispanic" ~ "hispanic",
    TRUE ~ variable
  ))|>
  inner_join(redlining_grades, by="variable")|>
  select(variable, A_avg:D_sd,  p.value)


baseline_char|>
  filter(!variable %in% c("obesity","diabetes","highchol"))|>
  knitr::kable(digits = 3)


baseline_char|>
  filter(variable %in% c("obesity","diabetes","highchol"))|>
  knitr::kable(digits = 3)
```

### Comparing census tracts with HRS for Grade A with Grade D

```{r warning=FALSE, message=FALSE}
q1_vs_q4 = 
final_analysis|>
  filter(red_grade %in%c("A","D"))|>
  mutate()

q1q4_desc = 
q1_vs_q4|>
  group_by(red_grade)|>
  summarise(
    ph_snap = mean(ph_snap),
    ph_with_children = mean(ph_with_children),
    ph_elderly = mean(ph_elderly),
    ph_below_poverty = mean(ph_below_poverty),
    ph_disability = mean(ph_disability),
    minority = mean(minority),
    ph_no_work = mean(ph_no_work),
    obesity = mean(obesity),
    highchol = mean(highchol),
    diabetes = mean(diabetes)
  )|>
 pivot_longer(cols = -red_grade, names_to = "variable", values_to = "value")|>
 pivot_wider(names_from = red_grade, names_prefix = "grade_", values_from = value)|>
 mutate(
    mean_diff = grade_D - grade_A,
  )|>
  rename(outcome = variable)
```


```{r }
uni_reg = function(dataset, outcomes, predictor) {
  results = list() 
  for (outcome in outcomes) {
    formula = as.formula(paste(outcome, "~", predictor))
    model= t.test(
      formula,
      data = dataset
    )
    results[[outcome]] = broom::tidy(model)
  }
  
  combined_results = bind_rows(results, .id = "outcome")
  return(combined_results)
}

outcomes = c("ph_snap","ph_with_children", "ph_elderly","ph_below_poverty", "ph_disability", "minority", "ph_no_work", "obesity", "highchol", "diabetes")


uni_reg(q1_vs_q4, outcomes, "red_grade")|>
  right_join(q1q4_desc, by="outcome")|>
  select(outcome, grade_A, grade_D, mean_diff, p.value)|>
  knitr::kable(digits = 3)
```

## Regression Analysis for Redlining and Food Insecurity

```{r}
# univariate analysis
lm(ph_snap~red_grade,final_analysis)|>
  broom::tidy()|>
  knitr::kable(digits= 3)

## multivariate model 1
red_snap_m1 = lm(ph_snap~red_grade+minority+ph_elderly+ph_disability+ph_no_work,final_analysis)
tidy_model = broom::tidy(red_snap_m1)
ci = confint(red_snap_m1)

tidy_model|>
  mutate(
    CI_Lower = ci[, 1],
    CI_Upper = ci[, 2],
  )|>
  select(term, estimate, p.value, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 1 Estimates with 95% Confidence Intervals")

## multivariate model 2
red_snap_m2 =
  final_analysis|>
  lm(log(ph_snap+1)~red_grade+minority+ph_elderly+ph_disability+ph_no_work, data=_)

tidy_model = broom::tidy(red_snap_m2)
ci = confint(red_snap_m2)

tidy_model|>
  mutate(
    Back_Transformed_Estimate = exp(estimate),
    CI_Lower = exp(ci[, 1]),
    CI_Upper = exp(ci[, 2]),
    p.value = ifelse(p.value<0.001, "<0.001", p.value)
  )|>
  select(term, estimate, p.value, Back_Transformed_Estimate, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 2 Coefficients with Back-Transformed Estimates and 95% Confidence Intervals")

##multivariate model 3
red_snap_m3 = lm(log(ph_snap+1)~red_grade+ph_white+ph_black+ph_asian+ph_elderly+ph_disability+ph_no_work,final_analysis)

tidy_model = broom::tidy(red_snap_m3)
ci = confint(red_snap_m3)

tidy_model|>
  mutate(
    Back_Transformed_Estimate = exp(estimate),
    CI_Lower = exp(ci[, 1]),
    CI_Upper = exp(ci[, 2]),
    p.value = ifelse(p.value<0.001, "<0.001", p.value)
  )|>
  select(term, estimate, p.value, Back_Transformed_Estimate, CI_Lower, CI_Upper)|>
  knitr::kable(digits = 3, caption = "Model 3 Coefficients with Back-Transformed Estimates and 95% Confidence Intervals")
```

### Residual Plots

```{r}
m1_residual = data.frame(
    fitted = fitted(red_snap_m1),
    residuals = residuals(red_snap_m1)
  ) 
m1_residual |>
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "deepskyblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values of Multivariate Model 1",
    x = "Fitted Values",
    y = "Residuals"
  )


m2_residual = data.frame(
    fitted = fitted(red_snap_m2),
    residuals = residuals(red_snap_m2)
  ) 
m2_residual |>
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "deepskyblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values of Multivariate Model 2",
    x = "Fitted Values",
    y = "Residuals"
  )


m3_residual = data.frame(
    fitted = fitted(red_snap_m3),
    residuals = residuals(red_snap_m3)
  ) 
m3_residual |>
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "deepskyblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values of Multivariate Model 3",
    x = "Fitted Values",
    y = "Residuals"
  )
```

### Cross Validation and Model Selection

```{r}
cv_df =
  crossv_mc(final_analysis, 100)
```


```{r}
cv_df =
cv_df|>
  mutate(model_1 = map(train, \(df)  
                       lm(ph_snap~red_grade+minority+ph_elderly+ph_disability+ph_no_work,
                          data=df)),
         model_2 = map(train, \(df) 
                       lm(log(ph_snap+1)~red_grade+minority+ph_elderly+ph_disability
                          +ph_no_work,data=df)),
         model_3 = map(train, \(df) 
                       lm(log(ph_snap+1)~red_grade+ph_white+ph_black+ph_asian+ph_elderly
                          +ph_disability+ph_no_work, data=df))
           )|>
  mutate(
    rmse_m1 = map2_dbl(model_1, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_m2 = map2_dbl(model_2, test, \(mod, df) rmse(model = mod, data = df)),
    rmse_m3 = map2_dbl(model_3, test, \(mod, df) rmse(model = mod, data = df))
  )
```

```{r}
cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  ggplot(aes(x = model, y = rmse)) + geom_violin()+
  labs(title = "Model 1 vs. Model 2 & 3")+
  theme_minimal()

cv_df |> 
  select(starts_with("rmse")) |> 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") |> 
  mutate(model = fct_inorder(model)) |> 
  filter(model == "m2" | model== "m3")|>
  ggplot(aes(x = model, y = rmse)) + geom_violin()+
  labs(title = "Model 2 vs. Model 3")+
  theme_minimal()
```

Based on the rootmean square models, model 1 and 2 perform signficantly better than model 1, while model 3 performs slightly better than model 3. 


## Statistical Analysis of Obesity 

### Does the relationship between redlining and health outcomes depend on minority percentage?

```{r}
uni_ob = lm(obesity ~ red_grade, data = final_analysis) 
uni_ob|> broom::tidy() |> knitr::kable(digits = 3)

mul_ob = lm(obesity ~ red_grade + minority, data = final_analysis) 
mul_ob |> broom::tidy() |> knitr::kable(digits = 3)

int_ob = lm(obesity ~ red_grade + minority + red_grade * minority, data = final_analysis) 
int_ob |> broom::tidy() |> knitr::kable(digits = 3)


uni_dia = lm(diabetes ~ red_grade, data = final_analysis)
uni_dia |> broom::tidy() |> knitr::kable(digits = 3)

mul_dia = lm(diabetes ~ red_grade + minority, data = final_analysis) 
mul_dia |> broom::tidy() |> knitr::kable(digits = 3)

int_dia = lm(diabetes ~ red_grade + minority + red_grade * minority, data = final_analysis) 
int_dia |> broom::tidy() |> knitr::kable(digits = 3)
```

### Model Comparisons

```{r}
AIC(uni_ob, mul_ob, int_ob)
models_ob = list(Univariate = uni_ob, Multivariate = mul_ob, Interaction = int_ob)
sapply(models_ob, function(model) summary(model)$adj.r.squared)

AIC(uni_dia, mul_dia, int_dia)
models_dia = list(Univariate = uni_dia, Multivariate = mul_dia, Interaction = int_dia)
sapply(models_dia, function(model) summary(model)$adj.r.squared)
```

### Logistic Regression 

```{r}
final_analysis <- final_analysis |> 
  mutate(high_obesity = ifelse(obesity > 30, 1, 0),
         high_diabetes = ifelse(diabetes > 10, 1, 0))

glm(high_obesity ~ red_grade + minority, data = final_analysis, family = binomial) |>
  broom::tidy() |> 
  mutate(odds_ratio = exp(estimate)) |>
  select(term, estimate, odds_ratio, statistic, std.error, p.value) |>
  knitr::kable(digits = 3)

glm(high_diabetes ~ red_grade + minority, data = final_analysis, family = binomial) |>
  broom::tidy() |> 
  mutate(odds_ratio = exp(estimate)) |>
  select(term, estimate, odds_ratio, statistic, std.error, p.value) |>
  knitr::kable(digits = 3)
```

### residual plots

```{r}
ob_residual = data.frame(
    fitted = fitted(int_ob),
    residuals = residuals(int_ob)
  ) 
ob_residual |>
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values of Obesity Regression Model",
    x = "Fitted Values",
    y = "Residuals"
  )

dia_residual = data.frame(
    fitted = fitted(int_dia),
    residuals = residuals(int_dia)
  ) 
dia_residual |>
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values of Diabetes Regression Model",
    x = "Fitted Values",
    y = "Residuals"
  )
```








