---
title: "Redlining in NYC"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(viridis)
library(shiny)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(stringr)
library(sf)
library(leaflet)
library(tidycensus)
```

<<<<<<< HEAD

```{r server, echo=FALSE}
server <- function(input, output, session) {
  data_cleaning_report <- renderUI({
    includeHTML("data_cleaning.html")
  })
}
=======
```{r datacleaning, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rmarkdown::render("data_cleaning_shiny.Rmd", quiet = TRUE)
>>>>>>> 8ffeb999a15a159e3fc275be7930dbd9e03eb635
```

```{r setup2, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
census_api_key("f7b3e47cd8e0fdf7f504751ffad10825be7a5ccf")
install = TRUE
nyc_tracts <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income as an example (this is just to pull the geometry)
  state = "NY",
  county = c("Bronx", "Kings", "New York", "Queens", "Richmond"),
  geometry = TRUE,
  year = 2020
) |>
  st_transform(crs = 4326)  # Transform to WGS84 for compatibility with Leaflet

nyc_tracts <- nyc_tracts |>
  mutate(borough = case_when(
    str_detect(NAME, "Bronx County") ~ "Bronx",
    str_detect(NAME, "Kings County") ~ "Brooklyn",
    str_detect(NAME, "New York County") ~ "Manhattan",
    str_detect(NAME, "Queens County") ~ "Queens",
    str_detect(NAME, "Richmond County") ~ "Staten Island",
    TRUE ~ "Unknown"  # This is a fallback in case there are other values
  ))
```

```{r, echo=FALSE, warning=FALSE}
nyc_tracts = nyc_tracts |>
  rename(geoid = GEOID)

redlining_sf = 
  nyc_tracts |>
  left_join(redlining_snap, by = "geoid")

redlining_sf <- redlining_sf |>
  mutate(
    borough = case_when(
      str_detect(NAME, "Bronx County") ~ "Bronx",
      str_detect(NAME, "Kings County") ~ "Brooklyn",
      str_detect(NAME, "New York County") ~ "Manhattan",
      str_detect(NAME, "Queens County") ~ "Queens",
      str_detect(NAME, "Richmond County") ~ "Staten Island",
      TRUE ~ "Unknown"
    )
  )

health_data =
  place_cleaned |>
  left_join(redlining_snap, by = "geoid") |>
  janitor::clean_names()

health_data_long <- health_data %>%
  drop_na(red_grade, obesity, diabetes, highchol) %>%
  pivot_longer(cols = c(obesity, diabetes, highchol), 
               names_to = "condition", values_to = "rate")

health_spatial_data <- nyc_tracts %>%
  left_join(health_data_long, by = "geoid")  # Make sure the 'geoid' column matches in both datasets
```



Column {.sidebar}
-----------------------------------------------------------------------

```{r datacleaning, echo=FALSE}
uiOutput("data_cleaning_report")
```

```{r, echo=FALSE}
health_indicator = 
	health_data_long |> 
	distinct(condition) |> 
	pull(condition)

radioButtons(
	inputId = "health_indicator", 
	label = h3("Health Indicator"), 
	choices = health_indicator,
	selected = "diabetes"
)
```


```{r, echo=FALSE}
checkboxInput(
  inputId = "store_check",
  label = h3("Retail Store"),
  value = TRUE
)
```


```{r, echo=FALSE}
borough_choices = 
	redlining_sf |> 
	distinct(borough) |> 
	pull(borough)

selectInput(
	inputId = "borough_choice", 
	label = h3("Select Borough"), 
	choices = borough_choices,
	selected = "Manhattan"
)
```




Column {data-width=650}
-----------------------------------------------------------------------

### Redlining Map

```{r, echo=FALSE}
renderLeaflet({
  filtered_data <- redlining_sf %>% filter(borough == input$borough_choice)

  # Create leaflet map with the filtered data
leaflet(filtered_data) |>
    addProviderTiles(providers$CartoDB.Positron) |> 
    addPolygons(
      fillColor = ~colorFactor(palette = c("#208e56", "#e4e44c", "#1595b4", "#ef647b"), domain = redlining_sf$red_grade)(red_grade),
      weight = 1,
      opacity = 1,
      color = "black",
      fillOpacity = 0.9,
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#666",
        fillOpacity = 0.9,
        bringToFront = TRUE
      ),
      label = ~paste0("Census Tract: ", geoid, "<br>Redlining Grade: ", red_grade)
    ) |>
    addLegend(
      pal = colorFactor(palette = c("#208e56", "#e4e44c", "#1595b4", "#ef647b"), domain = redlining_sf$red_grade),
      values = ~red_grade,
      opacity = 0.7,
      title = "Redlining Category",
      position = "bottomright"
    )
})
```

Column {data-width=350}
-----------------------------------------------------------------------

### Health Indicator Map

```{r, echo=FALSE}
renderLeaflet({
  req(input$health_indicator, input$borough_choice)
  filtered_data <- health_spatial_data %>%
    filter(condition == input$health_indicator & borough == input$borough_choice)

  leaflet(filtered_data) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%  
    addPolygons(
      fillColor = ~colorNumeric(palette = "YlOrRd", domain = health_spatial_data$rate)(rate),  
      weight = 1,
      opacity = 1,
      color = "black",
      fillOpacity = 0.7,
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#666",
        fillOpacity = 0.9,
        bringToFront = TRUE
      ),
      label = ~paste0("Condition: ", condition, "<br>Rate: ", round(rate, 2))
    ) %>%
    addLegend(
      pal = colorNumeric(palette = "YlOrRd", domain = health_spatial_data$rate),  
      values = ~rate,
      title = "Health Condition Rate",
      position = "bottomright"
    )
})
```


### Healthy Grocery Stores and Redlining

```{r, echo=FALSE}
renderLeaflet({
  filtered_redlining <- redlining_sf %>% filter(borough == input$borough_choice)
  
  map <- leaflet(filtered_redlining) |>
    addProviderTiles(providers$CartoDB.Positron) |>  
    addPolygons(
      fillColor = ~colorFactor(palette = c("#208e56", "#e4e44c", "#1595b4", "#ef647b"), domain = redlining_sf$red_grade)(red_grade),
      weight = 1,
      opacity = 1,
      color = "black",
      fillOpacity = 0.7,
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#666",
        fillOpacity = 0.9,
        bringToFront = TRUE
      ),
      label = ~paste0("Census Tract: ", geoid, 
                      "<br>Redlining Grade: ", red_grade,
                      "<br>Borough: ", borough)
    ) |>
    addLegend(
      position = "bottomright",
      pal = colorFactor(palette = c("#208e56", "#e4e44c", "#1595b4", "#ef647b"), domain = redlining_sf$red_grade),
      values = ~red_grade,
      title = "Redlining Grade",
      opacity = 0.7
    )
  
  if (input$store_check) {
    map <- map |>
      addCircleMarkers(
        data = nyc_healthy_store,
        lng = ~longitude,
        lat = ~latitude,
        radius = 5,
        color = "#0f6921",
        stroke = FALSE,
        fillOpacity = 0.7,
        label = ~paste0("Store Name: ", store_name,
                        "<br>Borough: ", borough,
                        "<br>ZIP Code: ", zip_code),
        labelOptions = labelOptions(
          noHide = FALSE,
          direction = "auto"
        )
      ) |>
      addLegend(
        position = "bottomleft",
        colors = "#0f6921",
        labels = "Healthy Grocery Store",
        title = "Healthy Grocery Stores in NYC",
        opacity = 0.7
      )
  }
  
  map
})
```




